---
title: "MA5832_Assignment_3_Kuruvilla_Rijo"
author: "Rijo"
date: "2023-10-08"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
# Load necessary libraries
library(caTools)
library(caret)
library(MASS)
library(corrplot)
library(tidyverse)
library(tidyr)
library(car)
library(ggplot2)
library(car)
library(summarytools)
library(lubridate)
# library(dplyr)
library(readxl)
library(psych)
library(e1071)
library(pROC)
library(glmnet)
library(rpart)
library(ipred)
#library(Hmisc)
library(mice)
library(missForest)
library(randomForest)


# # Neural Network Library
# # Installing Keras
# #install.packages("keras")
# library(keras)
# library(remotes)
# # remotes::install_github(paste0("rstudio/", c("reticulate", "tensorflow", "keras")))
# # reticulate::install_python()
# # install_tensorflow(envname = "r-tensorflow")
# library(tensorflow)

```

# Set the working directory
```{r}
getwd() # Get the current working directory
setwd('D:/DATA_SCIENCE_MASTERS/MA5832_Data_Mining_and_Machine_Learning/Assessment/A3') # set the working directory
```


# Load the Data Set
```{r}
unemployment_rate = read.csv("Aus_data_2023.csv")
# View(unemployment_rate) # View the data
str(unemployment_rate) # Structure of the data
dim(unemployment_rate)  # Dimensions of the data : 166 x 9 
```
 dim(unemployment_rate)
[1] 166   9

# EDA

# I. Change Column Names

# Original Column Names 
Y : unemployment rate measured in percentage
X1: Percentage change in Gross domestic product;
X2: Percentage change in the Government final consumption expenditure;
X3: Percentage change in final consumption expenditure of all industry sectors;
X4: Term of trade index (percentage)
X5: Consumer Price Index of all groups (CPI) ;
X6: Number of job vacancies measured in thousands;
X7: Estimated Resident Population in thousands

# Rename Columns
```{r}
# Set New Column Names
new_column_names = c("Year_Month","Unemployment_Rate","Percentage_change_GDP",
                     "Percentage_change_GFCE","Percentage_change_FCEAI","Term_of_TI",
                     "Consumer_Price_Index","Job_Vacancies","Estimated_Resident_Population")

# Rename the old column names to new column names
unemployment_rate = unemployment_rate %>% rename(Year_Month = "Year_Q",
                                  Unemployment_Rate = "Y",
                                  Percentage_change_GDP = "X1",
                                  Percentage_change_GFCE = "X2",
                                  Percentage_change_FCEAI = "X3",
                                  Term_of_TI = "X4",
                                  Consumer_Price_Index = "X5",
                                  Job_Vacancies = "X6",
                                  Estimated_Resident_Population = "X7")
# Renamed column names 
str(unemployment_rate)
```
# New Column Names
> str(unemployment_rate)

'data.frame':	166 obs. of  9 variables:
 $ Year_Month                   : chr  "1982_Dec" "1982_Jun" "1982_Mar" "1982_Sep" ...
 $ Unemployment_Rate            : num  8.79 6.57 6.21 7.11 9.71 ...
 $ Percentage_change_GDP        : num  -1.9 0.5 -1.2 -1 1.4 -0.5 -1.3 2.5 0.3 0.8 ...
 $ Percentage_change_GFCE       : num  5.3 7.9 -3.2 -5.9 5.5 2.1 4.8 -4.9 -11.4 3.7 ...
 $ Percentage_change_FCEAI      : num  1.7 3.4 -0.1 -1.3 0.5 -0.6 0.7 0.6 -0.5 -0.2 ...
 $ Term_of_TI                   : num  -1 2.5 -2.9 -0.5 1.3 1.2 0.5 -0.2 -0.3 0.5 ...
 $ Consumer_Price_Index         : num  33.6 31.5 30.8 32.6 36.5 35 34.3 35.6 37.4 36.4 ...
 $ Job_Vacancies                : num  28.6 31.2 36.4 28.4 37.6 31.7 30.3 34.3 52.3 43.5 ...
 $ Estimated_Resident_Population: num  15289 15184 15122 15239 15484 ...
 
---------------------------------------------------------------------------------------------

---------------------------------------------------------------------------------------------
# Data Type Transformations 
```{r}
# 1. Transform the "Year_Q" column
str(unemployment_rate)

# Create a new data frame with the Year_Month column
year_month = data.frame(Year_Month = unemployment_rate$Year_Month)

# Split the Year_Month column into Year and Month Column
year_month = separate(year_month, Year_Month,into = c("Year","Month"), sep = "_")
#str(year_month)

# Add this new column to the Original Data Frame
unemployment_rate = cbind(unemployment_rate,year_month)
# Remove the first column which consists of the "Year_Month" combined data
unemployment_rate = unemployment_rate[,-1]

#-------------------------------------------------------------------------------------------
# Convert  "Year" to "Date" Format

unemployment_rate$Year = as.Date(unemployment_rate$Year , format('%Y'))
# Extract only Year
unemployment_rate$Year = year(unemployment_rate$Year)
str(unemployment_rate)

#------------------------------------------------------------------------------------------
# Create a new column named "Quarter"

months = c("Mar", "Jun", "Sep", "Dec")
quarters = c(1, 2, 3, 4)
# Create the "Quarter" column based on the "Month" values
unemployment_rate$Quarter = quarters[match(unemployment_rate$Month, months)]
#View(unemployment_rate)
str(unemployment_rate)
```


# 1. Check for Missing Values
```{r}
is.na(unemployment_rate)  # Check for missing values
colSums(is.na(unemployment_rate))  # Count the number of missing values per column
# which(colSums(is.na(unemployment_rate))>0) # Identify the position of the columns with atleast one missing value
names(which(colSums(is.na(unemployment_rate))>0))  # Return the column names with missing values
```

> colSums(is.na(unemployment_rate))  # Count the number of missing values per column

            Unemployment_Rate         Percentage_change_GDP        Percentage_change_GFCE       Percentage_change_FCEAI 
                            0                             1                             1                             1 
                   Term_of_TI          Consumer_Price_Index                 Job_Vacancies Estimated_Resident_Population 
                            1                             0                             5                             2 
                         Year                         Month                       Quarter 
                            0                             0                             0
# Comments: The following are the columns with the number of missing values 
"Percentage_change_GDP" = 1, 
"Percentage_change_GFCE" = 1, 
"Percentage_change_FCEAI" = 1,
"Term_of_TI" = 1, 
"Job_Vacancies" = 5
"Estimated_Resident_Population" = 2 

## -- i. Using Uivariate Imputation to Handle Missing Values
```{r}

## BEFORE IMPUTATION
# Check if the variables are  normally distributed or not, before proceeding with mean or median
summary(unemployment_rate$Percentage_change_GDP)
summary(unemployment_rate$Percentage_change_GFCE)
summary(unemployment_rate$Percentage_change_FCEAI)
summary(unemployment_rate$Term_of_TI)
summary(unemployment_rate$Job_Vacancies)
summary(unemployment_rate$Estimated_Resident_Population)

# Transforming the Left Skewed Vairable 
summary(unemployment_rate$Percentage_change_GFCE)
summary(sqrt(unemployment_rate$Percentage_change_GFCE)) # Square- Root Transformation
summary((unemployment_rate$Percentage_change_GFCE)^(1/3)) # Cube- Root Transformation

# GGPLOT RESULTS - Before Imputation
p1 = ggplot(data = unemployment_rate, mapping = aes(Percentage_change_GDP)) + geom_histogram(bins = 40, colour = "black") + ggtitle('Percentage_change_GDP - Before Imputation')
p2 = ggplot(data = unemployment_rate, mapping = aes(Percentage_change_GFCE)) + geom_histogram(bins = 40, colour = "black") + ggtitle('Percentage_change_GFCE - Before Imputation')
p3 = ggplot(data = unemployment_rate, mapping = aes(Percentage_change_FCEAI)) + geom_histogram(bins = 40, colour = "black") + ggtitle('Percentage_change_FCEAI - Before Imputation')
p4 = ggplot(data = unemployment_rate, mapping = aes(Term_of_TI)) + geom_histogram(bins = 40, colour = "black") + ggtitle('Term_of_TI - Before Imputation')
p5 = ggplot(data = unemployment_rate, mapping = aes(Job_Vacancies)) + geom_histogram(bins = 40, colour = "black") + ggtitle('Job_Vacancies - Before Imputation')
p6 = ggplot(data = unemployment_rate, mapping = aes(Estimated_Resident_Population)) + geom_histogram(bins = 40, colour = "black") + ggtitle('Estimated_Resident_Population - Before Imputation')
gridExtra::grid.arrange(p1,p2,p3,p4,p5,p6, nrow = 3)

# Apply Univariate Imputation using MEDIAN
imputed_Percentage_change_GDP = replace(unemployment_rate$Percentage_change_GDP,
                     is.na(unemployment_rate$Percentage_change_GDP),
                     median(unemployment_rate$Percentage_change_GDP , na.rm = TRUE))
imputed_Percentage_change_GFCE = replace(unemployment_rate$Percentage_change_GFCE,
                     is.na(unemployment_rate$Percentage_change_GFCE),
                     median(unemployment_rate$Percentage_change_GFCE , na.rm = TRUE))
imputed_Percentage_change_FCEAI = replace(unemployment_rate$Percentage_change_FCEAI,
                     is.na(unemployment_rate$Percentage_change_FCEAI),
                     median(unemployment_rate$Percentage_change_FCEAI , na.rm = TRUE))
imputed_Term_of_TI = replace(unemployment_rate$Term_of_TI,
                     is.na(unemployment_rate$Term_of_TI),
                     median(unemployment_rate$Term_of_TI , na.rm = TRUE))
imputed_Job_Vacancies = replace(unemployment_rate$Job_Vacancies,
                     is.na(unemployment_rate$Job_Vacancies),
                     median(unemployment_rate$Job_Vacancies , na.rm = TRUE))
imputed_Estimated_Resident_Population = replace(unemployment_rate$Estimated_Resident_Population,
                     is.na(unemployment_rate$Estimated_Resident_Population),
                     median(unemployment_rate$Estimated_Resident_Population , na.rm = TRUE))

# GGPLOT RESULTS - After Median Imputation 
p1_1 = ggplot(as.data.frame(imputed_Percentage_change_GDP), mapping = aes(imputed_Percentage_change_GDP)) + geom_histogram(bins = 40, colour = "pink") + ggtitle('Percentage_change_GDP - After Median Imputation')
p1_2 = ggplot(as.data.frame(imputed_Percentage_change_GFCE), mapping = aes(imputed_Percentage_change_GFCE)) + geom_histogram(bins = 40, colour = "pink") + ggtitle('Percentage_change_GFCE - After Median Imputation')
p1_3 = ggplot(as.data.frame(imputed_Percentage_change_FCEAI), mapping = aes(imputed_Percentage_change_FCEAI)) + geom_histogram(bins = 40, colour = "pink") + ggtitle('Percentage_change_FCEAI - After Median Imputation')
p1_4 = ggplot(as.data.frame(imputed_Term_of_TI), mapping = aes(imputed_Term_of_TI)) + geom_histogram(bins = 40, colour = "pink") + ggtitle('Term_of_TI - After Median Imputation')
p1_5 = ggplot(as.data.frame(imputed_Job_Vacancies), mapping = aes(imputed_Job_Vacancies)) + geom_histogram(bins = 40, colour = "pink") + ggtitle('Job_Vacancies - After Median Imputation')
p1_6 = ggplot(as.data.frame(imputed_Estimated_Resident_Population), mapping = aes(imputed_Estimated_Resident_Population)) + geom_histogram(bins = 40, colour = "pink") + ggtitle('Estimated_Resident_Population - After Median Imputation')
gridExtra::grid.arrange(p1_1,p1_2,p1_3,p1_4,p1_5,p1_6, nrow = 3)

```
                
# -- ii.  Missing Values using Predictive Mean Matching (PMM)
```{r}
unemployment_rate_1 = unemployment_rate
set.seed(99) # set seed to maintain model consistency
# Impute missing values using PMM
pmm_imputed_Percentage_change_GDP_1 = complete(mice(unemployment_rate_1, method = "pmm"))$Percentage_change_GDP
pmm_imputed_Percentage_change_GFCE_1 = complete(mice(unemployment_rate_1, method = "pmm"))$Percentage_change_GFCE 
pmm_imputed_Percentage_change_FCEAI_1 = complete(mice(unemployment_rate_1, method = "pmm"))$Percentage_change_FCEAI
pmm_imputed_Term_of_TI_1 = complete(mice(unemployment_rate_1, method = "pmm"))$Term_of_TI
pmm_imputed_Job_Vacancies_1 = complete(mice(unemployment_rate_1, method = "pmm"))$Job_Vacancies
pmm_imputed_Estimated_Resident_Population_1 = complete(mice(unemployment_rate_1, method = "pmm"))$Estimated_Resident_Population
    
pmm_imputed_values = data.frame(pmm_imputed_Percentage_change_GDP_1,
              pmm_imputed_Percentage_change_GFCE_1,
              pmm_imputed_Percentage_change_FCEAI_1,
              pmm_imputed_Term_of_TI_1,
              pmm_imputed_Job_Vacancies_1,
              pmm_imputed_Estimated_Resident_Population_1)
# View(imputed_values)

is.na(pmm_imputed_values)  # Check for missing values
colSums(is.na(pmm_imputed_values))  # Count the number of missing values per column
## Missing Values successfully imputed using PMM


# Checking the distribution of the data after imputation

p1_1 =  ggplot(data = pmm_imputed_values, mapping = aes(pmm_imputed_Percentage_change_GDP_1)) + geom_histogram(bins = 40, colour = "green") + ggtitle('Percentage_change_GDP - After Imputation using PMM')

p1_2 =  ggplot(data = pmm_imputed_values, mapping = aes(pmm_imputed_Percentage_change_GFCE_1)) + geom_histogram(bins = 40, colour = "green") + ggtitle('Percentage_change_GFCE - After Imputation using PMM')

p1_3 =  ggplot(data = pmm_imputed_values, mapping = aes(pmm_imputed_Percentage_change_FCEAI_1)) + geom_histogram(bins = 40, colour = "green") + ggtitle('Percentage_change_FCEAI - After Imputation using PMM')

p1_4 =  ggplot(data = pmm_imputed_values, mapping = aes(pmm_imputed_Term_of_TI_1)) + geom_histogram(bins = 40, colour = "green") + ggtitle('Term_of_TI - After Imputation using PMM')

p1_5 =  ggplot(data = pmm_imputed_values, mapping = aes(pmm_imputed_Job_Vacancies_1)) + geom_histogram(bins = 40, colour = "green") + ggtitle('Job_Vacancies - After Imputation using PMM')

p1_6 =  ggplot(data = pmm_imputed_values, mapping = aes(pmm_imputed_Estimated_Resident_Population_1)) + geom_histogram(bins = 40, colour = "green") + ggtitle('Estimated_Resident_Population - After Imputation using PMM')

gridExtra::grid.arrange(p1_1,p1_2,p1_3,p1_4,p1_5,p1_6, nrow = 3)

# After Imputation Results

## BEFORE IMPUTATION
# Check if the variables are  normally distributed or not, before proceeding with mean or median
summary(pmm_imputed_values$imputed_Percentage_change_GDP_1)
summary(pmm_imputed_values$imputed_Percentage_change_GFCE_1)
summary(pmm_imputed_values$imputed_Percentage_change_FCEAI_1)
summary(pmm_imputed_values$imputed_Term_of_TI_1)
summary(pmm_imputed_values$imputed_Job_Vacancies_1)
summary(pmm_imputed_values$imputed_Estimated_Resident_Population_1)
```

Before Imputation:
> summary(unemployment_rate$Percentage_change_GDP)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
 -6.800   0.000   0.400   0.403   0.800   3.800       1 
> summary(unemployment_rate$Percentage_change_GFCE)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
-11.400  -0.100   1.400   1.175   2.600  14.900       1 
> summary(unemployment_rate$Percentage_change_FCEAI)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
-8.2000  0.4000  0.8000  0.7867  1.2000  6.0000       1 
> summary(unemployment_rate$Term_of_TI)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
-9.7000 -1.1000  0.4000  0.4927  2.0000 11.3000       1 
> summary(unemployment_rate$Job_Vacancies)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   28.4    70.4   105.1   132.2   173.6   454.2       5 
> summary(unemployment_rate$Estimated_Resident_Population)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
  15122   17469   19522   20102   22857   26268       2

After Imputation:
> summary(imputed_values$imputed_Percentage_change_GDP_1)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-6.8000  0.0000  0.4000  0.4042  0.7750  3.8000 
> summary(imputed_values$imputed_Percentage_change_GFCE_1)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-11.400  -0.175   1.400   1.138   2.600  14.900 
> summary(imputed_values$imputed_Percentage_change_FCEAI_1)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-8.2000  0.4000  0.7500  0.7801  1.2000  6.0000 
> summary(imputed_values$imputed_Term_of_TI_1)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-9.7000 -1.0750  0.4000  0.5139  2.0750 11.3000 
> summary(imputed_values$imputed_Job_Vacancies_1)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.40   70.85  108.05  133.24  173.57  454.20 
> summary(imputed_values$imputed_Estimated_Resident_Population_1)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  15122   17490   19577   20171   23014   26268 

-----------------------------------------------------------------------------

#iii. ---Imputed Values Using CART

```{r}

# Compare the Distributions
unemployment_rate_2 = unemployment_rate
set.seed(99) # set seed to maintain model consistency
# Impute missing values using PMM
cart_imputed_Percentage_change_GDP_1 = complete(mice(unemployment_rate_2, method = "cart"))$Percentage_change_GDP
cart_imputed_Percentage_change_GFCE_1 = complete(mice(unemployment_rate_2, method = "cart"))$Percentage_change_GFCE 
cart_imputed_Percentage_change_FCEAI_1 = complete(mice(unemployment_rate_2, method = "cart"))$Percentage_change_FCEAI
cart_imputed_Term_of_TI_1 = complete(mice(unemployment_rate_2, method = "cart"))$Term_of_TI
cart_imputed_Job_Vacancies_1 = complete(mice(unemployment_rate_2, method = "cart"))$Job_Vacancies
cart_imputed_Estimated_Resident_Population_1 = complete(mice(unemployment_rate_2, method = "cart"))$Estimated_Resident_Population
    
cart_imputed_values = data.frame(cart_imputed_Percentage_change_GDP_1,
              cart_imputed_Percentage_change_GFCE_1,
              cart_imputed_Percentage_change_FCEAI_1,
              cart_imputed_Term_of_TI_1,
              cart_imputed_Job_Vacancies_1,
              cart_imputed_Estimated_Resident_Population_1)
# View(cart_imputed_values)

# Checking the distribution of the data after imputation - CART

p1_1_cart =  ggplot(data = cart_imputed_values, mapping = aes(cart_imputed_Percentage_change_GDP_1)) + geom_histogram(bins = 40, colour = "yellow") + ggtitle('Percentage_change_GDP - After Imputation using CART')

p1_2_cart =  ggplot(data = cart_imputed_values, mapping = aes(cart_imputed_Percentage_change_GFCE_1)) + geom_histogram(bins = 40, colour = "yellow") + ggtitle('Percentage_change_GFCE - After Imputation using CART')

p1_3_cart =  ggplot(data = cart_imputed_values, mapping = aes(cart_imputed_Percentage_change_FCEAI_1)) + geom_histogram(bins = 40, colour = "yellow") + ggtitle('Percentage_change_FCEAI - After Imputation using CART')

p1_4_cart = ggplot(data = cart_imputed_values, mapping = aes(cart_imputed_Term_of_TI_1)) + geom_histogram(bins = 40, colour = "yellow") + ggtitle('Term_of_TI - After Imputation using CART')

p1_5_cart =  ggplot(data = cart_imputed_values, mapping = aes(cart_imputed_Job_Vacancies_1)) + geom_histogram(bins = 40, colour = "yellow") + ggtitle('Job_Vacancies - After Imputation using CART')

p1_6_cart =  ggplot(data = cart_imputed_values, mapping = aes(cart_imputed_Estimated_Resident_Population_1)) + geom_histogram(bins = 40, colour = "yellow") + ggtitle('Estimated_Resident_Population - After Imputation using CART')

gridExtra::grid.arrange(p1_1_cart,p1_2_cart,p1_3_cart,p1_4_cart,p1_5_cart,p1_6_cart, nrow = 3)


```

------------------------------------------------------------------------------

#iii. ---Imputed Values Using LASSO

```{r}
# Compare the Distributions
unemployment_rate_3 = unemployment_rate
set.seed(99) # set seed to maintain model consistency
# Impute missing values using PMM
lasso_imputed_Percentage_change_GDP_1 = complete(mice(unemployment_rate_3, method = "lasso.norm"))$Percentage_change_GDP
lasso_imputed_Percentage_change_GFCE_1 = complete(mice(unemployment_rate_3, method = "lasso.norm"))$Percentage_change_GFCE 
lasso_imputed_Percentage_change_FCEAI_1 = complete(mice(unemployment_rate_3, method = "lasso.norm"))$Percentage_change_FCEAI
lasso_imputed_Term_of_TI_1 = complete(mice(unemployment_rate_3, method = "lasso.norm"))$Term_of_TI
lasso_imputed_Job_Vacancies_1 = complete(mice(unemployment_rate_3, method = "lasso.norm"))$Job_Vacancies
lasso_imputed_Estimated_Resident_Population_1 = complete(mice(unemployment_rate_3, method = "lasso.norm"))$Estimated_Resident_Population
    
lasso_imputed_values = data.frame(lasso_imputed_Percentage_change_GDP_1,
              lasso_imputed_Percentage_change_GFCE_1,
              lasso_imputed_Percentage_change_FCEAI_1,
              lasso_imputed_Term_of_TI_1,
              lasso_imputed_Job_Vacancies_1,
              lasso_imputed_Estimated_Resident_Population_1)
# View(lasso_imputed_values)

# Checking the distribution of the data after imputation - CART

p1_1_lasso =  ggplot(data = lasso_imputed_values, mapping = aes(lasso_imputed_Percentage_change_GDP_1)) + geom_histogram(bins = 40, colour = "blue") + ggtitle('Percentage_change_GDP - After Imputation using LASSO')

p1_2_lasso =  ggplot(data = lasso_imputed_values, mapping = aes(lasso_imputed_Percentage_change_GFCE_1)) + geom_histogram(bins = 40, colour = "blue") + ggtitle('Percentage_change_GFCE - After Imputation using LASSO')

p1_3_lasso =  ggplot(data = lasso_imputed_values, mapping = aes(lasso_imputed_Percentage_change_FCEAI_1)) + geom_histogram(bins = 40, colour = "blue") + ggtitle('Percentage_change_FCEAI - After Imputation using LASSO')

p1_4_lasso = ggplot(data = lasso_imputed_values, mapping = aes(lasso_imputed_Term_of_TI_1)) + geom_histogram(bins = 40, colour = "blue") + ggtitle('Term_of_TI - After Imputation using LASSO')

p1_5_lasso =  ggplot(data = lasso_imputed_values, mapping = aes(lasso_imputed_Job_Vacancies_1)) + geom_histogram(bins = 40, colour = "blue") + ggtitle('Job_Vacancies - After Imputation using LASSO')

p1_6_lasso = ggplot(data = lasso_imputed_values, mapping = aes(lasso_imputed_Estimated_Resident_Population_1)) + geom_histogram(bins = 40, colour = "blue") + ggtitle('Estimated_Resident_Population - After Imputation using LASSO')

gridExtra::grid.arrange(p1_1_lasso,p1_2_lasso,p1_3_lasso,p1_4_lasso,p1_5_lasso,p1_6_lasso, nrow = 3)

########################################################################################
cbind(pmm_imputed_values,cart_imputed_values)
View(head(cbind(pmm_imputed_values,cart_imputed_values),5))
```


------------------------------------------------------------------------------

#iii. ---Imputed Values Using Random Forrest

```{r}

# Compare the Distributions
unemployment_rate_4 = unemployment_rate
unemployment_rate_4 = unemployment_rate_4[,-c(10)]
str(unemployment_rate_4)

set.seed(99) # set seed to maintain model consistency
# Impute missing values using PMM
rf_imputed_Percentage_change_GDP_1 = missForest(unemployment_rate_4)$ximp$Percentage_change_GDP
rf_imputed_Percentage_change_GFCE_1 = missForest(unemployment_rate_4)$ximp$Percentage_change_GFCE
rf_imputed_Percentage_change_FCEAI_1 = missForest(unemployment_rate_4)$ximp$Percentage_change_FCEAI
rf_imputed_Term_of_TI_1 = missForest(unemployment_rate_4)$ximp$Term_of_TI
rf_imputed_Job_Vacancies_1 = missForest(unemployment_rate_4)$ximp$Job_Vacancies
rf_imputed_Estimated_Resident_Population_1 = missForest(unemployment_rate_4)$ximp$Estimated_Resident_Population
    
rf_imputed_values = data.frame(rf_imputed_Percentage_change_GDP_1,
              rf_imputed_Percentage_change_GFCE_1,
              rf_imputed_Percentage_change_FCEAI_1,
              rf_imputed_Term_of_TI_1,
              rf_imputed_Job_Vacancies_1,
              rf_imputed_Estimated_Resident_Population_1)
# View(rf_imputed_values)


#####################################

p1_1_rf =  ggplot(data = rf_imputed_values, mapping = aes(rf_imputed_Percentage_change_GDP_1)) + geom_histogram(bins = 40, colour = "cyan") + ggtitle('Percentage_change_GDP - After Imputation using RF')

p1_2_rf =  ggplot(data = rf_imputed_values, mapping = aes(rf_imputed_Percentage_change_GFCE_1)) + geom_histogram(bins = 40, colour = "cyan") + ggtitle('Percentage_change_GFCE - After Imputation using RF')

p1_3_rf =  ggplot(data = rf_imputed_values, mapping = aes(rf_imputed_Percentage_change_FCEAI_1)) + geom_histogram(bins = 40, colour = "cyan") + ggtitle('Percentage_change_FCEAI - After Imputation using RF')

p1_4_rf = ggplot(data = rf_imputed_values, mapping = aes(rf_imputed_Term_of_TI_1)) + geom_histogram(bins = 40, colour = "cyan") + ggtitle('Term_of_TI - After Imputation using RF')

p1_5_rf =  ggplot(data = rf_imputed_values, mapping = aes(rf_imputed_Job_Vacancies_1)) + geom_histogram(bins = 40, colour = "cyan") + ggtitle('Job_Vacancies - After Imputation using RF')

p1_6_rf = ggplot(data = rf_imputed_values, mapping = aes(rf_imputed_Estimated_Resident_Population_1)) + geom_histogram(bins = 40, colour = "cyan") + ggtitle('Estimated_Resident_Population - After Imputation using RF')

gridExtra::grid.arrange(p1_1_rf,p1_2_rf,p1_3_rf,p1_4_rf,p1_5_rf,p1_6_rf, nrow = 3)

# cbind(pmm_imputed_values[,c(1)],cart_imputed_values[,c(1)],lasso_imputed_values[,c(1)],rf_imputed_values[,c(1)])
# View(cbind(pmm_imputed_values[,c(1)],cart_imputed_values[,c(1)],lasso_imputed_values[,c(1)],rf_imputed_values[,c(1)]))


```

# Histograms ( Distributions) after NA imputations

```{r}
###################
#
p1 = ggplot(data = unemployment_rate, mapping = aes(Percentage_change_GDP)) + geom_histogram(bins = 40, colour = "black") + ggtitle('Percentage_change_GDP - Before Imputation') # Original 

p1_1_pmm =  ggplot(data = pmm_imputed_values, mapping = aes(pmm_imputed_Percentage_change_GDP_1)) + geom_histogram(bins = 40, colour = "green") + ggtitle('Percentage_change_GDP - After Imputation using PMM')

p1_1_cart =  ggplot(data = cart_imputed_values, mapping = aes(cart_imputed_Percentage_change_GDP_1)) + geom_histogram(bins = 40, colour = "yellow") + ggtitle('Percentage_change_GDP - After Imputation using CART')

p1_1_lasso =  ggplot(data = lasso_imputed_values, mapping = aes(lasso_imputed_Percentage_change_GDP_1)) + geom_histogram(bins = 40, colour = "blue") + ggtitle('Percentage_change_GDP - After Imputation using LASSO')
gridExtra::grid.arrange(p1_1_pmm,p1_1_cart,p1_1_lasso,nrow = 1)

p1_1_rf =  ggplot(data = rf_imputed_values, mapping = aes(rf_imputed_Percentage_change_GDP_1)) + geom_histogram(bins = 40, colour = "blue") + ggtitle('Percentage_change_GDP - After Imputation using RF')
gridExtra::grid.arrange(p1,p1_1_pmm,p1_1_cart,p1_1_lasso,p1_1_rf,nrow = 3)

cbind(pmm_imputed_values[,c(1)],cart_imputed_values[,c(1)],lasso_imputed_values[,c(1)],rf_imputed_values[,c(1)])
View(cbind(pmm_imputed_values[,c(1)],cart_imputed_values[,c(1)],lasso_imputed_values[,c(1)],rf_imputed_values[,c(1)]))
```

------------------------------------------------------------------------------
# Restructure the Data
```{r}
# pmm_imputed_values
# cart_imputed_values
# lasso_imputed_values
rf_imputed_values

# Add these values to the original data set
str(unemployment_rate)
ur_au = cbind(unemployment_rate,rf_imputed_values) # Combine the 2 data frames
str(ur_au)

# Filter the original columns which have missing values
#str(ur_au[,-c(2,3,4,5,7,8)])
ur_au= ur_au[,-c(2,3,4,5,7,8)]
str(ur_au)

# Rename the columns 
new_column_names_2 = c("Percentage_change_GDP","Percentage_change_GFCE","Percentage_change_FCEAI","Term_of_TI","Job_Vacancies","Estimated_Resident_Population") # Set New Column Names

# Rename the old column names to new column names
ur_au = ur_au %>% rename(Percentage_change_GDP = "rf_imputed_Percentage_change_GDP_1",
                   Percentage_change_GFCE = "rf_imputed_Percentage_change_GFCE_1",
                   Percentage_change_FCEAI = "rf_imputed_Percentage_change_FCEAI_1",
                   Term_of_TI = "rf_imputed_Term_of_TI_1",
                   Job_Vacancies = "rf_imputed_Job_Vacancies_1",
                   Estimated_Resident_Population = "rf_imputed_Estimated_Resident_Population_1")
# Renamed column names 
ur_au_rearranged = ur_au[,c("Unemployment_Rate",
                            "Year",
                            "Month",
                            "Quarter",
                            "Percentage_change_GDP",
                            "Percentage_change_GFCE",
                            "Percentage_change_FCEAI",
                            "Term_of_TI",
                            "Consumer_Price_Index",
                            "Job_Vacancies",
                            "Estimated_Resident_Population")]

# str(ur_au_rearranged)

unemployment_rate = ur_au_rearranged  # assign the rearranged columns to the original data
#unemployment_rate = ur_au_rearranged[,-c(2,3,4)] # Remove Year, Month and Quarter since it's not a part of the original predictors
#str(unemployment_rate)

which(is.na(unemployment_rate))  # Check for missing values
colSums(is.na(unemployment_rate))  # Count the number of missing values per column


```
Original DS:                                                                                                                
> str(unemployment_rate)
'data.frame':	166 obs. of  9 variables:
 $ Year_Month                   : chr  "1982_Dec" "1982_Jun" "1982_Mar" "1982_Sep" ...
 $ Unemployment_Rate            : num  8.79 6.57 6.21 7.11 9.71 ...
 $ Percentage_change_GDP        : num  -1.9 0.5 -1.2 -1 1.4 -0.5 -1.3 2.5 0.3 0.8 ...
 $ Percentage_change_GFCE       : num  5.3 7.9 -3.2 -5.9 5.5 2.1 4.8 -4.9 -11.4 3.7 ...
 $ Percentage_change_FCEAI      : num  1.7 3.4 -0.1 -1.3 0.5 -0.6 0.7 0.6 -0.5 -0.2 ...
 $ Term_of_TI                   : num  -1 2.5 -2.9 -0.5 1.3 1.2 0.5 -0.2 -0.3 0.5 ...
 $ Consumer_Price_Index         : num  33.6 31.5 30.8 32.6 36.5 35 34.3 35.6 37.4 36.4 ...
 $ Job_Vacancies                : num  28.6 31.2 36.4 28.4 37.6 31.7 30.3 34.3 52.3 43.5 ...
 $ Estimated_Resident_Population: num  15289 15184 15122 15239 15484 ...

After Transformation:
> str(unemployment_rate)
'data.frame':	166 obs. of  11 variables:
 $ Unemployment_Rate            : num  8.79 6.57 6.21 7.11 9.71 ...
 $ Year                         : num  1982 1982 1982 1982 1983 ...
 $ Month                        : chr  "Dec" "Jun" "Mar" "Sep" ...
 $ Quarter                      : num  4 2 1 3 4 2 1 3 4 2 ...
 $ Percentage_change_GDP        : num  -1.9 0.5 -1.2 -1 1.4 -0.5 -1.3 2.5 0.3 0.8 ...
 $ Percentage_change_GFCE       : num  5.3 7.9 -3.2 -5.9 5.5 2.1 4.8 -4.9 -11.4 3.7 ...
 $ Percentage_change_FCEAI      : num  1.7 3.4 -0.1 -1.3 0.5 -0.6 0.7 0.6 -0.5 -0.2 ...
 $ Term_of_TI                   : num  -1 2.5 -2.9 -0.5 1.3 1.2 0.5 -0.2 -0.3 0.5 ...
 $ Consumer_Price_Index         : num  33.6 31.5 30.8 32.6 36.5 35 34.3 35.6 37.4 36.4 ...
 $ Job_Vacancies                : num  28.6 31.2 36.4 28.4 37.6 31.7 30.3 34.3 52.3 43.5 ...
 $ Estimated_Resident_Population: num  15289 15184 15122 15239 15484 ..


# Further EDA
- Correlation Plots
- Outlier Analysis
```{r}

# Numerical Variable Analysis

# 1. Using Pair Plot
#psych::pairs.panels(unemployment_rate[,-c(3)], main = "Unemployment Rate  - Pair Plot Analysis")
psych::pairs.panels(unemployment_rate, main = "Unemployment Rate  - Pair Plot Analysis")

# 2. Using Correlation Plot
unemployment_rate_cor = cor(unemployment_rate[,-c(3)])
#unemployment_rate_cor = cor(unemployment_rate)
corrplot::corrplot(unemployment_rate_cor, method = "number",title = "Correlation Plot - Numerical Variables")

# Scatter Plot Analysis - Predictors vs Response

# ------------------------------------------------------------------------------------------#
# 1. Year vs Unemployment Rate
sc_1 = ggplot(data = unemployment_rate, aes(x = Year, y = Unemployment_Rate)) +
  geom_point(color = "red") +
  labs(x = "Year", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Year vs Unemployment_Rate")


# sc_l_line = ggplot(data = unemployment_rate, aes(x = Year, y = Unemployment_Rate)) +
#   geom_smooth(method = "auto", color = "blue") +  # Add a smoothed line
#   labs(x = "Year", y = "Unemployment Rate") +
#   ggtitle("Scatter Plot Analysis - Year vs Unemployment_Rate")

# ------------------------------------------------------------------------------------------#
# 2. Percentage_change_GDP vs Unemployment Rate
sc_2 = ggplot(data = unemployment_rate, aes(x = Percentage_change_GDP , y = Unemployment_Rate)) +
  geom_point(color = "green") +
  labs(x = "Percentage_change_GDP", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - % change in Gross Domestic Product vs Unemployment_Rate")

# ------------------------------------------------------------------------------------------#
# 3. Percentage_change_GFCE vs Unemployment Rate
sc_3 = ggplot(data = unemployment_rate, aes(x = Percentage_change_GFCE , y = Unemployment_Rate)) +
  geom_point(color = "blue") +
  labs(x = "Percentage_change_GFCE", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - % change in  Government final consumption expenditure vs Unemployment_Rate")

# ------------------------------------------------------------------------------------------#
# 4. Percentage_change_FCEAI vs Unemployment Rate
sc_4 = ggplot(data = unemployment_rate, aes(x = Percentage_change_FCEAI , y = Unemployment_Rate)) +
  geom_point(color = "purple") +
  labs(x = "Percentage_change_FCEAI", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - % change in  final consumption expenditure of all industry sectors vs Unemployment_Rate")

# ------------------------------------------------------------------------------------------#
# 5. Term_of_TI vs Unemployment Rate
sc_5 = ggplot(data = unemployment_rate, aes(x = Term_of_TI , y = Unemployment_Rate)) +
  geom_point(color = "orange") +
  labs(x = " Term of trade index", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Term_of_TI vs Unemployment_Rate")

# ------------------------------------------------------------------------------------------#
# 6. Consumer_Price_Index vs Unemployment Rate
sc_6 = ggplot(data = unemployment_rate, aes(x = Consumer_Price_Index , y = Unemployment_Rate)) +
  geom_point(color = "magenta") +
  labs(x = "Consumer_Price_Index", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Consumer Price Index of all groups vs Unemployment_Rate")

# ------------------------------------------------------------------------------------------#
# 7. Job_Vacancies vs Unemployment Rate
sc_7 = ggplot(data = unemployment_rate, aes(x = Job_Vacancies , y = Unemployment_Rate)) +
  geom_point(color = "navy") +
  labs(x = "Job_Vacancies", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Job Vacancies vs Unemployment_Rate")

# ------------------------------------------------------------------------------------------#
# 8. Estimated_Resident_Population vs Unemployment Rate
sc_8 = ggplot(data = unemployment_rate, aes(x = Estimated_Resident_Population , y = Unemployment_Rate)) +
  geom_point(color = "turquoise") +
  labs(x = "Estimated_Resident_Population", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Resident Population (in Thousands) vs Unemployment_Rate")


gridExtra::grid.arrange(sc_1, sc_2, sc_3, sc_4 , sc_5 , sc_6, sc_7 , sc_8, nrow = 2)
```

# Smoothened Line Plots
```{r}
# 1. Year vs Unemployment Rate
sc_1 = ggplot(data = unemployment_rate, aes(x = Year, y = Unemployment_Rate)) +
  geom_point(color = "red") +
  labs(x = "Year", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Year vs Unemployment_Rate")

sc_l_line = ggplot(data = unemployment_rate, aes(x = Year, y = Unemployment_Rate)) +
  geom_smooth(method = "auto", color = "blue") +  # Add a smoothed line
  labs(x = "Year", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Year vs Unemployment_Rate")
gridExtra::grid.arrange(sc_1, sc_l_line, nrow = 2)

# MAX
(max_value = max(unemployment_rate$Unemployment_Rate)) # To determine Peak Unemployment Rate # 11.14877
(year_at_max = unemployment_rate$Year[which(unemployment_rate$Unemployment_Rate == max_value)]) # To determine Year # 1992

# MIN
(min_value = min(unemployment_rate$Unemployment_Rate)) # To determine Peak Unemployment Rate # 3.472797
(year_at_min = unemployment_rate$Year[which(unemployment_rate$Unemployment_Rate == min_value)]) # To determine Year # 2022


# Eyeball test: Check the unemployment rate between 2006 and 2010
(check_2006_2010 = unemployment_rate[unemployment_rate$Year >= 2006 & unemployment_rate$Year <= 2010, ])
(min_unemployment_rate = min(check_2006_2010$Unemployment_Rate)) # 4.086032  # Find the minimum unemployment rate in the filtered data
(year_at_min_2 = check_2006_2010$Year[which(check_2006_2010$Unemployment_Rate == min_unemployment_rate)]) # to check the exact year  # 2008

gridExtra::grid.arrange(sc_1, sc_l_line, nrow = 2)
---------------------------------
# SMOOTH - LINE PLOTS

# 1. Year vs Unemployment Rate
sl_1 = ggplot(data = unemployment_rate, aes(x = Year, y = Unemployment_Rate)) +
  labs(x = "Year", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Year vs Unemployment_Rate") +
  geom_smooth(method = "auto", color = "blue")  # Add a smoothed line

# 2. Percentage_change_GDP vs Unemployment Rate
sl_2 = ggplot(data = unemployment_rate, aes(x = Percentage_change_GDP , y = Unemployment_Rate)) +
  labs(x = "Percentage_change_GDP", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - % change in Gross Domestic Product vs Unemployment_Rate") +
  geom_smooth(method = "auto", color = "green")  # Add a smoothed line

# 3. Percentage_change_GFCE vs Unemployment Rate
sl_3 = ggplot(data = unemployment_rate, aes(x = Percentage_change_GFCE , y = Unemployment_Rate)) +
  labs(x = "Percentage_change_GFCE", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - % change in  Government final consumption expenditure vs Unemployment_Rate") +
  geom_smooth(method = "auto", color = "blue")  # Add a smoothed line

# 4. Percentage_change_FCEAI vs Unemployment Rate
sl_4 = ggplot(data = unemployment_rate, aes(x = Percentage_change_FCEAI , y = Unemployment_Rate)) +
  labs(x = "Percentage_change_FCEAI", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - % change in  final consumption expenditure of all industry sectors vs Unemployment_Rate") +
  geom_smooth(method = "auto", color = "purple")  # Add a smoothed line

# gridExtra::grid.arrange(sl_1, sl_2, sl_3,sl_4, nrow = 2)

# 5. Term_of_TI vs Unemployment Rate
sl_5 = ggplot(data = unemployment_rate, aes(x = Term_of_TI , y = Unemployment_Rate)) +
  labs(x = " Term of trade index", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Term_of_TI vs Unemployment_Rate") +
  geom_smooth(method = "auto", color = "orange")  # Add a smoothed line

# 6. Consumer_Price_Index vs Unemployment Rate
sl_6 = ggplot(data = unemployment_rate, aes(x = Consumer_Price_Index , y = Unemployment_Rate)) +
  labs(x = "Consumer_Price_Index", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Consumer Price Index of all groups vs Unemployment_Rate") +
  geom_smooth(method = "auto", color = "magenta")  # Add a smoothed line

# 7. Job_Vacancies vs Unemployment Rate
sl_7 = ggplot(data = unemployment_rate, aes(x = Job_Vacancies , y = Unemployment_Rate)) +
  labs(x = "Job_Vacancies", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Job Vacancies vs Unemployment_Rate") +
  geom_smooth(method = "auto", color = "navy")  # Add a smoothed line

# 8. Estimated_Resident_Population vs Unemployment Rate
sl_8 = ggplot(data = unemployment_rate, aes(x = Estimated_Resident_Population , y = Unemployment_Rate)) +
  labs(x = "Estimated_Resident_Population", y = "Unemployment Rate") +
  ggtitle("Scatter Plot Analysis - Resident Population (in Thousands) vs Unemployment_Rate") +
  geom_smooth(method = "auto", color = "turquoise")  # Add a smoothed line

gridExtra::grid.arrange(sl_1, sl_2, sl_3,sl_4,sl_5, sl_6,sl_7,sl_8 ,nrow = 2)

```


# Descriptive Statistics& Box Plots
```{r}
summary(unemployment_rate$Year)
summary(unemployment_rate$Unemployment_Rate)
summary(unemployment_rate$Percentage_change_GDP)
summary(unemployment_rate$Percentage_change_GFCE)
summary(unemployment_rate$Percentage_change_FCEAI)
summary(unemployment_rate$Term_of_TI)
summary(unemployment_rate$Consumer_Price_Index)
summary(unemployment_rate$Job_Vacancies)
summary(unemployment_rate$Estimated_Resident_Population)

# Boxplot for Unemployment Rate
gg_1 = ggplot(unemployment_rate, aes(y = Unemployment_Rate)) +
  geom_boxplot(fill = "blue", color = "black") +
  labs(title = "Boxplot of Unemployment Rate", y = "Unemployment Rate")

# Boxplot for Percentage Change in GDP
gg_2 = ggplot(unemployment_rate, aes(y = Percentage_change_GDP)) +
  geom_boxplot(fill = "green", color = "black") +
  labs(title = "Boxplot of Percentage Change in GDP", y = "Percentage Change in GDP")

# Boxplot for Percentage Change in GFCE
gg_3 = ggplot(unemployment_rate, aes(y = Percentage_change_GFCE)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(title = "Boxplot of Percentage Change in GFCE", y = "Percentage Change in GFCE")

# Boxplot for Percentage Change in FCEAI
gg_4 = ggplot(unemployment_rate, aes(y = Percentage_change_FCEAI)) +
  geom_boxplot(fill = "purple", color = "black") +
  labs(title = "Boxplot of Percentage Change in FCEAI", y = "Percentage Change in FCEAI")

# gridExtra::grid.arrange(gg_1, gg_2, gg_3, gg_4 , nrow = 2)

# Boxplot for Consumer Price Index
gg_5 = ggplot(unemployment_rate, aes(y = Term_of_TI)) +
  geom_boxplot(fill = "red", color = "black") +
  labs(title = "Boxplot of Consumer Price Index", y = "Consumer Price Index")

# Boxplot for Consumer Price Index
gg_6 = ggplot(unemployment_rate, aes(y = Consumer_Price_Index)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(title = "Boxplot of Consumer Price Index", y = "Consumer Price Index")

# Boxplot for Job Vacancies
gg_7= ggplot(unemployment_rate, aes(y = Job_Vacancies)) +
  geom_boxplot(fill = "brown", color = "black") +
  labs(title = "Boxplot of Job Vacancies", y = "Job Vacancies")

# Boxplot for Estimated Resident Population
gg_8 = ggplot(unemployment_rate, aes(y = Estimated_Resident_Population)) +
  geom_boxplot(fill = "pink", color = "black") +
  labs(title = "Boxplot of Estimated Resident Population", y = "Estimated Resident Population")

gridExtra::grid.arrange(gg_1, gg_2, gg_3, gg_4 ,gg_5, gg_6, gg_7,gg_8, nrow = 2)

```

> summary(unemployment_rate$Unemployment_Rate)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  3.473   5.279   6.187   6.714   8.172  11.149 
> summary(unemployment_rate$Percentage_change_GDP)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-6.8000  0.0000  0.4000  0.4007  0.7750  3.8000 
> summary(unemployment_rate$Percentage_change_GFCE)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-11.400  -0.100   1.400   1.175   2.600  14.900 
> summary(unemployment_rate$Percentage_change_FCEAI)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-8.2000  0.4000  0.7500  0.7854  1.2000  6.0000 
> summary(unemployment_rate$Term_of_TI)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-9.7000 -1.0750  0.4000  0.5032  2.0750 11.3000 
> summary(unemployment_rate$Consumer_Price_Index)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  30.80   59.90   77.35   79.19  102.30  133.70 
> summary(unemployment_rate$Job_Vacancies)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.40   70.85  108.05  133.52  174.20  454.20 
> summary(unemployment_rate$Estimated_Resident_Population)
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  15122   17490   19577   20173   23014   26268 
  
# ----------------------------------------------------------------------------------------------
# Data Partioning
```{r}
# Data Partitioning 
set.seed(99) # Set the seed for model consistency 

# Training Data = Dec 1982 to Dec 2020
ur_training_filter = (unemployment_rate$Year < 2021) | (unemployment_rate$Year == 2020 & unemployment_rate$Month == "Dec")

# Test Data = Jan 2021 to March 2023 
ur_test_filter = (unemployment_rate$Year >= 2021) | (unemployment_rate$Year == 2021 & unemployment_rate$Month %in% c("Mar"))

# Create the training and test sets
ur_train = unemployment_rate[ur_training_filter, ]
ur_test = unemployment_rate[ur_test_filter, ]

dim(ur_train)  #  156 x  11
dim(ur_test)  #  10 x 11

ur_train = ur_train[,-c(2,3,4)] # Remove Year, Month and Quarter since it's not a part of the original predictors
ur_test = ur_test[,-c(2,3,4)] # Remove Year, Month and Quarter since it's not a part of the original predictors

#View(ur_train)
#View(ur_test)

dim(ur_train)  #  156 x 8
dim(ur_test)  #  10 x 8
```

# ---------------------------------------------- ML MODELS --------------------------------------------------------


# ---------------------------------------------- SHRINKAGE METHODS --------------------------------------------------------

# 1. RIDGE REGRESSION
```{r}

# Build a matrix of Predictors
x = model.matrix(Unemployment_Rate ~ . , data = ur_train)[,-1] # remove the response variable
y = ur_train$Unemployment_Rate  # response variable

# Build the Ridge Regression Model on Training Data
set.seed(99)
ridge_ur = glmnet(x,y, family = 'gaussian', aplha = 0)
ridge_ur
summary(ridge_ur) # Summary of the model
plot(ridge_ur) # Plot the model

# ------------------------------- CROSS VALIDATION using a grid of Lambda values------------------------# 

# Set the lambda values
grid = 10^seq(10, -2 , length = 100)
 
# Applying Cross-validation
start_time_1 = Sys.time()  # Start the timer

set.seed(99)

ridge_model_cv = cv.glmnet(x,y, alpha = 0, family = 'gaussian', lambda = grid, scale = TRUE) # Build the 10 - fold CV Model on training data

end_time_1 = Sys.time()  # End the timer 
(time_diff_1 = end_time_1 - start_time_1) # Time Difference = 0.109123 secs


ridge_model_cv  # Model Results 
summary(ridge_model_cv) # Model Summary

# Plot the model for the CV_Lasso
plot(ridge_model_cv, xvar = "lambda", main = "Ridge CV using a grid of Lambda Values") 

# Comments:
# The plot displays the Mean-Squared Error (MSE) according to the log of lambda. The left dashed vertical line indicates that the log of the optimal value of lambda is approximately -2, which is the one that minimizes the prediction error. This lambda value will give the most accurate model. The exact value of lambda can be viewed as follow:

# Minimal Lambda
(best_lambda_ridge = ridge_model_cv$lambda.min)# Minimum Lambda = 0.01

log10(0.01) # -2  # Comments: Which confirms the value displayed on the graph.


#Coefficients : Using the lambda.min on 10-fold CV
coef(ridge_model_cv, ridge_model_cv$lambda.min)
```
> coef(ridge_model_cv, ridge_model_cv$lambda.min)
8 x 1 sparse Matrix of class "dgCMatrix"
                                         s1
(Intercept)                    1.0195451130
Percentage_change_GDP          0.4224458498
Percentage_change_GFCE         0.0669639897
Percentage_change_FCEAI       -0.3155397703
Term_of_TI                    -0.0213959833
Consumer_Price_Index          -0.0599233820
Job_Vacancies                 -0.0356059158
Estimated_Resident_Population  0.0007361817


# ----------------------------------------X---------------------------------------------------------------# 
# Build the Ridge Model using the Best Lambda Value (0.01)
```{r}

start_time_2 = Sys.time()  # Start the timer
set.seed(99)
ridge_best_model = glmnet(x, y, alpha = 0, lambda = best_lambda_ridge, scale = TRUE) # Build the model using the Best Lambda value

end_time_2 = Sys.time()  # End the timer 
(time_diff_2 = end_time_2 - start_time_2) # Time Difference = 0.005417109  secs

ridge_best_model # Model results
summary(ridge_best_model) # Model summary

ridge_best_model$lambda  # 0.01

coef(ridge_best_model)  # Print the Coefficients


# ------------------------------ PERFORMANCE ON TRAINING DATA  ------------------------------------#
# Prediction on Training Data using Best (Minimum Lambda)
ridge_train_predict = predict(ridge_best_model, s = ridge_best_model$lambda, newx = x)

# Mean Squared Error for Training Data
(mse_ridge_train = mean((ridge_train_predict - y)^2))

# Root Mean Squared Error for Training Data
(rmse_ridge_train = sqrt(mse_ridge_train))

# Report Training Performance Metrics
cat("Training Performance Metrics:\n")
cat("MSE on Training Data: ", mse_ridge_train, "\n")  # 0.8782203
cat("RMSE on Training Data: ", rmse_ridge_train, "\n") # 0.9371341 

# ------------------------------ PERFORMANCE ON TEST DATA  ------------------------------------#
# Prediction on Test Data using Best (Minimum Lambda)
x_test = model.matrix(Unemployment_Rate ~ ., data = ur_test)[, -1]

ridge_test_predict = predict(ridge_best_model, newx = x_test)  # Predictions on the Test Data

# Mean Squared Error for Test Data
(mse_ridge_test = mean((ridge_test_predict - ur_test$Unemployment_Rate)^2))

# Root Mean Squared Error for Test Data
(rmse_ridge_test = sqrt(mse_ridge_test))

# Report Test Performance Metrics
cat("\nTest Performance Metrics:\n")
cat("MSE on Test Data: ", mse_ridge_test, "\n") # 37.40976 
cat("RMSE on Test Data: ", rmse_ridge_test, "\n") # 6.116352 

```
> ridge_best_model

Call:  glmnet(x = x, y = y, alpha = 0, lambda = best_lambda_ridge, scale = TRUE) 

  Df %Dev Lambda
1  7 72.5   0.01

> coef(ridge_best_model) # Print the Coeffecients
8 x 1 sparse Matrix of class "dgCMatrix"
                                         s0
(Intercept)                    1.0193081079
Percentage_change_GDP          0.4224455357
Percentage_change_GFCE         0.0669633879
Percentage_change_FCEAI       -0.3155382155
Term_of_TI                    -0.0213952666
Consumer_Price_Index          -0.0599256065
Job_Vacancies                 -0.0356061558
Estimated_Resident_Population  0.0007362036

> # Report Training Performance Metrics
> cat("Training Performance Metrics:\n")
Training Performance Metrics:
> cat("MSE on Training Data: ", mse_ridge_train, "\n")  # 0.8782203
MSE on Training Data:  0.8782203 
> cat("RMSE on Training Data: ", rmse_ridge_train, "\n") # 0.9371341 
RMSE on Training Data:  0.9371341 
> # Report Test Performance Metrics
> cat("\nTest Performance Metrics:\n")

Test Performance Metrics:
> cat("MSE on Test Data: ", mse_ridge_test, "\n") # 37.40976 
MSE on Test Data:  37.40976 
> cat("RMSE on Test Data: ", rmse_ridge_test, "\n") # 6.116352 
RMSE on Test Data:  6.116352 

# ------------------------------------------------ LASSO REGRESSION ------------------------------------------------------#
# 2. LASSO REGRESSION

```{r}
# Build a matrix of Predictors
x = model.matrix(Unemployment_Rate ~ . , data = ur_train)[,-1] # remove the response variable
y = ur_train$Unemployment_Rate # Response Variable


# Build the Lasso Regression Model on Training Data
lasso_ur = glmnet(x,y, family = 'gaussian', aplha = 1)
lasso_ur
summary(lasso_ur) # Summary of the model
plot(lasso_ur) # Plot the model

# ------------------------------- CROSS VALIDATION using a grid of Lambda values------------------------# 
# Set the lambda values
grid = 10^seq(10, -2 , length = 100)

# Applying Cross-validation
start_time_3 = Sys.time()  # Start the timer

set.seed(99)  # Set seed for model consistency
lasso_model_cv = cv.glmnet(x,y, alpha = 1, family = 'gaussian', lambda = grid, scale = TRUE) # Build the model using the training data

end_time_3 = Sys.time()  # End the timer 
(time_diff_3 = end_time_3 - start_time_3)  # Time difference of 0.1514041 secs

lasso_model_cv # Model results
summary(lasso_model_cv) # Model summary

# Plot the model for the CV_Lasso
plot(lasso_model_cv, main = "Lasso CV using a grid of Lambda Values") 

# Comments:
# The plot displays the Mean-Squared Error (MSE) according to the log of lambda. The left dashed vertical line indicates that the log of the optimal value of lambda is approximately -2, which is the one that minimizes the prediction error. This lambda value will give the most accurate model. The exact value of lambda can be viewed as follow:

# Minimal Lambda
(best_lambda_lasso = lasso_model_cv$lambda.min)# Minimum Lambda = 0.01

log10(0.01) # -2  # Comments: Which confirms the value displayed on the graph.


#Coefficients : Using the lambda.min on 10-fold CV
coef(lasso_model_cv, lasso_model_cv$lambda.min)
```
> #Coefficients : Using the lambda.min on 10-fold CV
> coef(lasso_model_cv, lasso_model_cv$lambda.min)
8 x 1 sparse Matrix of class "dgCMatrix"
                                         s1
(Intercept)                    0.4939532205
Percentage_change_GDP          0.3910860932
Percentage_change_GFCE         0.0577757850
Percentage_change_FCEAI       -0.2808455828
Term_of_TI                    -0.0157995655
Consumer_Price_Index          -0.0635513561
Job_Vacancies                 -0.0364967755
Estimated_Resident_Population  0.0007816274

# Build the Lasso Model using the Best Lambda Value (0.01)
```{r}
start_time_4 = Sys.time()  # Start the timer

set.seed(99)
lasso_best_model = glmnet(x,y, alpha = 1, lambda = best_lambda_lasso,scale = TRUE)

end_time_4 = Sys.time()  # End the timer 
(time_diff_4 = end_time_4 - start_time_4) # Time difference of 0.005409002 secs


lasso_best_model # Model results
lasso_best_model$lambda  # 0.01

coef(lasso_best_model) # Print the Coeffecients

#c = coef(lasso_best_model) # Print the Coeffecients
# ------------------------------ PERFORMANCE ON TRAINING DATA  ------------------------------------#
# Prediction on Training Data using Best (Minimum Lambda)
lasso_train_predict = predict(lasso_best_model, s = lasso_best_model$lambda, newx = x)

# Mean Squared Error for Training Data
(mse_lasso_train = mean((lasso_train_predict - y)^2))

# Root Mean Squared Error for Training Data
(rmse_lasso_train = sqrt(mse_lasso_train))

# Report Training Performance Metrics
cat("Training Performance Metrics:\n")
cat("MSE on Training Data: ", mse_lasso_train, "\n")  # 0.8782203
cat("RMSE on Training Data: ", rmse_lasso_train, "\n") # 0.9371341 

# ------------------------------ PERFORMANCE ON TEST DATA  ------------------------------------#

# Prediction on Test Data using Best (Minimum Lambda)
x_test = model.matrix(Unemployment_Rate ~ ., data = ur_test)[, -1]

lasso_test_predict = predict(lasso_best_model, newx = x_test)  # Predictions on the Test Data

# Mean Squared Error for Test Data
(mse_lasso_test = mean((lasso_test_predict - ur_test$Unemployment_Rate)^2))

# Root Mean Squared Error for Test Data
(rmse_lasso_test = sqrt(mse_lasso_test))

# Report Test Performance Metrics
cat("\nTest Performance Metrics:\n")
cat("MSE on Test Data: ", mse_lasso_test, "\n") 
cat("RMSE on Test Data: ", rmse_lasso_test, "\n") 

```
> lasso_best_model

Call:  glmnet(x = x, y = y, alpha = 1, lambda = best_lambda_lasso) 

  Df  %Dev Lambda
1  7 72.65   0.01

# Beta Values
> coef(lasso_best_model) # Print the Coeffecients
8 x 1 sparse Matrix of class "dgCMatrix"
                                        s0
(Intercept)                    0.493861056
Percentage_change_GDP          0.391090780
Percentage_change_GFCE         0.057777988
Percentage_change_FCEAI       -0.280843523
Term_of_TI                    -0.015800676
Consumer_Price_Index          -0.063553110
Job_Vacancies                 -0.036496531
Estimated_Resident_Population  0.000781637

> # Report Training Performance Metrics
> cat("Training Performance Metrics:\n")
Training Performance Metrics:
> cat("MSE on Training Data: ", mse_lasso_train, "\n")  # 0.8782203
MSE on Training Data:  0.8734861 
> cat("RMSE on Training Data: ", rmse_lasso_train, "\n") # 0.9371341 
RMSE on Training Data:  0.9346048 
> # Report Test Performance Metrics
> cat("\nTest Performance Metrics:\n")

Test Performance Metrics:
> cat("MSE on Test Data: ", mse_lasso_test, "\n") 
MSE on Test Data:  39.24328 
> cat("RMSE on Test Data: ", rmse_lasso_test, "\n") 
RMSE on Test Data:  6.264446 

# ---------------------------------------------------x------------------------------------------------------#

# -------------------------------------# II. Tree Based Models----------------------------------------------#


# Fit CART Regression Tree to the Data
```{r}
start_time_5 = Sys.time()  # Start the timer

set.seed(99) # Set seed for model consistency
tree_ur = rpart(Unemployment_Rate ~. , data = ur_train) # Build the CART model on training data 

end_time_5 = Sys.time()  # End the timer 
(time_diff_5 = end_time_5 - start_time_5) # Time difference of 0.08120894 secs

summary(tree_ur) # Model summary

# Plot the model
par(xpd = TRUE)
rattle::fancyRpartPlot(tree_ur, main = "Basic CART") # using rattle

# Bar Plot to see variable importance
barplot(tree_ur$variable.importance, cex.names  = 0.8, las = 2, , col = "red") 

# Predictions on test dat
#help("predict.rpart")
tree_ur_predict = predict(tree_ur , ur_test)

# Plot the graph
#plot(tree_ur_predict, ur_test$Unemployment_Rate)

# Mean Squared Error
mean((tree_ur_predict - ur_test$Unemployment_Rate)^2)  # 2.950931

# Plot the Complexity Parameter
par(mfrow = c(1,1))
plotcp(tree_ur, main = "CART - Complexity Parameter")  # cp = 0.021

# --------------------------------------------USING CV = 10 ------------------------------------------------------------#
start_time_5 = Sys.time()  # Start the timer

cv_tree_ur = rpart(Unemployment_Rate ~. , data = ur_train, xval = 10, model = TRUE) # Build the model using 10-FOld CV on training data

end_time_5 = Sys.time()  # End the timer 
(time_diff_5 = end_time_5 - start_time_5) # Time difference of 0.008260965 secs

summary(cv_tree_ur) # Model summary

rpart.plot::rpart.plot(cv_tree_ur, yesno = TRUE , main = "CART with CV") # Plot the tree

varImp(cv_tree_ur) # Variable Importance
plot(cv_tree_ur$variable.importance) # Variable Importance plot

# Plot the Complexity Parameter
plotcp(cv_tree_ur) #  cp = 0.021

# --------------------------------------------------------------------------------------------------------#
start_time_6 = Sys.time()  # Start the timer

# Build the tree model  using the Complexity Parameter
tree_ur_prune = prune(cv_tree_ur, cp = 0.021) 

end_time_6 = Sys.time()  # End the timer 
(time_diff_6 = end_time_6 - start_time_6) # Time difference of 0.00465703 sec

# Plot the Tree
rattle::fancyRpartPlot(tree_ur_prune, main = "CART with CP ") # using rattle  # After Pruning

# Bar Plot to see variable importance AFTER PRUNE
barplot(tree_ur_prune$variable.importance, cex.names  = 0.8, las = 2 , col = "green") 

# -------------------------- PERFORMANCE ON TRAINING DATA ---------------------------------#
# Predictions on Training Data
tree_ur_prune_predict_train = predict(tree_ur_prune, ur_train)

# Mean Squared Error after Pruning
(mse_cart_train = mean((tree_ur_prune_predict_train - ur_train$Unemployment_Rate)^2))

# Root Mean Squared Error after Pruning
(rmse_cart_train = sqrt(mse_cart_train)) 


# -------------------------- PERFORMANCE ON TEST DATA --------------------------------------#
# Predictions on Test Data
tree_ur_prune_predict_test = predict(tree_ur_prune , ur_test)

# Plot the graph
#plot(tree_ur_predict, ur_test$Unemployment_Rate)

# Mean Squared Error after Pruning
(mse_cart_test = mean((tree_ur_prune_predict_test - ur_test$Unemployment_Rate)^2)) #  1.901695

# Roor Mean Squared Error after Pruning
(rmse_cart_test = sqrt(mse_cart_test)) # 1.37902

#---------------------------- COMBINED RESULTS -------------------------------------------#
# Report Training Performance Metrics
cat("Training Performance Metrics:\n")
cat("MSE on Training Data: ", mse_cart_train, "\n")  
cat("RMSE on Training Data: ", rmse_cart_train, "\n") 

# Report Test Performance Metrics
cat("\nTest Performance Metrics:\n")
cat("MSE on Test Data: ", mse_cart_test, "\n") 
cat("RMSE on Test Data: ", rmse_cart_test, "\n") 

# ----------------------------COLLECTIVE PLOTS-------------------------------------------# 
par(mfrow = c(1,2))
barplot(tree_ur$variable.importance, cex.names  = 0.8, las = 2, , col = "red", main = "Before Pruning") 
barplot(tree_ur_prune$variable.importance, cex.names  = 0.8, las = 2 , col = "green", , main = "After Pruning") 
par(mfrow = c(1, 1))

```
> tree_ur_prune
n= 156 

node), split, n, deviance, yval
      * denotes terminal node

 1) root 156 498.2579000  6.873081  
   2) Estimated_Resident_Population>=18791.4 87  44.5126500  5.582670  
     4) Consumer_Price_Index>=78.85 70  25.6429300  5.364633 *
     5) Consumer_Price_Index< 78.85 17   1.8391650  6.480469 *
   3) Estimated_Resident_Population< 18791.4 69 126.2153000  8.500121  
     6) Job_Vacancies>=59.35 42  33.9454000  7.891349  
      12) Estimated_Resident_Population< 17435.3 22  16.5834900  7.445358  
        24) Consumer_Price_Index>=49.75 8   0.9718164  6.373717 *
        25) Consumer_Price_Index< 49.75 14   1.1744420  8.057725 *
      13) Estimated_Resident_Population>=17435.3 20   8.1723750  8.381939 *
     7) Job_Vacancies< 59.35 27  52.4917900  9.447100  
      14) Estimated_Resident_Population< 17260.7 15  23.5158700  8.559156 *
      15) Estimated_Resident_Population>=17260.7 12   2.3659270 10.557030 *
------------------------------------------
> # Report Training Performance Metrics
> cat("Training Performance Metrics:\n")
Training Performance Metrics:
> cat("MSE on Training Data: ", mse_cart_train, "\n")  
MSE on Training Data:  0.4082213 
> cat("RMSE on Training Data: ", rmse_cart_train, "\n") 
RMSE on Training Data:  0.638922 
> 
> # Report Test Performance Metrics
> cat("\nTest Performance Metrics:\n")

Test Performance Metrics:
> cat("MSE on Test Data: ", mse_cart_test, "\n") 
MSE on Test Data:  1.901695 
> cat("RMSE on Test Data: ", rmse_cart_test, "\n") 
RMSE on Test Data:  1.37902 

# --------------------------------------- BAGGING ------------------------------------------------------------#
```{r}

# Data Partitioning
set.seed(99) # Set the seed for model consistency

# Training Data = Dec 1982 to Dec 2020
ur_training_filter = (unemployment_rate$Year < 2021) | (unemployment_rate$Year == 2020 & unemployment_rate$Month == "Dec")
# Test Data = Jan 2021 to March 2023
ur_test_filter = (unemployment_rate$Year >= 2021) | (unemployment_rate$Year == 2021 & unemployment_rate$Month %in% c("Mar"))# Data Partitioning
set.seed(99) # Set the seed for model consistency

# Training Data = Dec 1982 to Dec 2020
ur_training_filter = (unemployment_rate$Year < 2021) | (unemployment_rate$Year == 2020 & unemployment_rate$Month == "Dec")
# Test Data = Jan 2021 to March 2023
ur_test_filter = (unemployment_rate$Year >= 2021) | (unemployment_rate$Year == 2021 & unemployment_rate$Month %in% c("Mar"))

# Create the training and test sets
ur_train = unemployment_rate[ur_training_filter, ]
ur_test = unemployment_rate[ur_test_filter, ]

# Remove Year, Month, and Quarter since it's not part of the original predictors
ur_train = ur_train[, -c(2, 3, 4)]
ur_test = ur_test[, -c(2, 3, 4)]

# Create the training and test sets
ur_train = unemployment_rate[ur_training_filter, ]
ur_test = unemployment_rate[ur_test_filter, ]

# Remove Year, Month, and Quarter since it's not part of the original predictors
ur_train = ur_train[, -c(2, 3, 4)]
ur_test = ur_test[, -c(2, 3, 4)]

#----------------------------------------------------------------------------------
start_time_7 = Sys.time()  # Start the timer

set.seed(99) # Set the seed for model consistency
# Hyperparameters for Bagging
hyperparameters = expand.grid(ntree = c(20:500)) # Vary the number of trees

# Initialize variables to store the best MSE and corresponding hyperparameters for training data
best_mse_train = Inf 
best_hyperparameters_train = NULL

# Bagging loop
for (i in 1:nrow(hyperparameters)) 
  {
    ntree = hyperparameters$ntree[i]
  
    # Create a bootstrapped sample from the training data
    bootstrap_sample = ur_train[sample(1:nrow(ur_train), replace = TRUE), ]
  
    # Train a random forest regression model (bagging) with the current hyperparameters
    ur_bag_cart = randomForest(Unemployment_Rate ~. ,
                             data = bootstrap_sample,
                             ntree = ntree,
                             mtry = 7,
                             scale = TRUE,
                             importance = TRUE)

    # Make predictions on the training set
    predictions_train = predict(ur_bag_cart, newdata = ur_train)
  
    # Calculate MSE for the training set
    mse_train = mean((predictions_train - ur_train$Unemployment_Rate)^2)

    # Check if this model has a better MSE compared to the overall best for training data
    if (mse_train < best_mse_train) 
      {
        best_mse_train = mse_train
        best_hyperparameters_train = ntree
        best_model_bag_train = ur_bag_cart
      }
}

end_time_7 = Sys.time()  # End the timer
(time_diff_7 = end_time_7 - start_time_7) # Time difference of 57.03038 secs

# Calculate RMSE for training data
(rmse_train = sqrt(best_mse_train))

# # Print the best MSE for training data
# cat("Best Mean Squared Error (MSE) for Training Data:", best_mse_train, "\n")
# 
# # Print the RMSE for training data
# cat("Root Mean Squared Error (RMSE) for Training Data:", rmse_train, "\n")

# Print the best hyperparameters for training data
cat("Best Hyperparameters for Training Data:\n",
    "Number of Trees (ntree):", best_hyperparameters_train, "\n")

print(best_model_bag_train) # Print the Best model obtained from the Bagging

best_model_bag_train$importance # Variable Importance 

# Variable Importance Plot
varImpPlot(best_model_bag_train, main = "Variable Importance - Bagging")
```
> cat("Best Hyperparameters for Training Data:\n",
+     "Number of Trees (ntree):", best_hyperparameters_train, "\n")
Best Hyperparameters for Training Data:
 Number of Trees (ntree): 134 
> 
> print(best_model_bag_train)

Call:
 randomForest(formula = Unemployment_Rate ~ ., data = bootstrap_sample,      ntree = ntree, mtry = 7, scale = TRUE, importance = TRUE) 
               Type of random forest: regression
                     Number of trees: 134
No. of variables tried at each split: 7

          Mean of squared residuals: 0.1183736
                    % Var explained: 96.63

> (time_diff_7 = end_time_7 - start_time_7) # Time difference of 0.00465703 sec
Time difference of 57.03038 secs

> best_model_bag_train$importance
                                 %IncMSE IncNodePurity
Percentage_change_GDP         0.10159991     10.647916
Percentage_change_GFCE        0.04742719      8.127313
Percentage_change_FCEAI       0.02393412      3.745323
Term_of_TI                    0.04934037      5.363021
Consumer_Price_Index          1.59638792    140.257095
Job_Vacancies                 2.26139278    142.720687
Estimated_Resident_Population 2.56916218    230.364305               

# Build The model using the Best Hyperparameters

```{r}
# Train the final model with the best hyperparameters on the full training data
start_time_8 = Sys.time()  # Start the timer
set.seed(99)
final_bagged_model = randomForest(Unemployment_Rate ~. , 
                              data = ur_train, 
                              mtry = 7,
                              ntree = 134,
                              importance = TRUE,
                              scale = TRUE) # using all 7 predictors

#start_time_8 = Sys.time()  # Start the timer
end_time_8 = Sys.time()  # End the timer
(time_diff_8 = end_time_8 - start_time_8) # Time difference of 0.07677102 secs

final_bagged_model # Model results 
final_bagged_model$importance  # Variable Importance

# Variable Importance Plot
varImpPlot(final_bagged_model, main = "Variable Importance - Bagging with the Optimal Hyperparameters")

# -------------------------- PERFORMANCE ON TRAINING DATA ---------------------------------#
# Prediction on Training Data
predicted_bag_trees_train = predict(final_bagged_model, ur_train)

cat("Training Performance Metrics:\n")
# MSE
(mse_bag_train = mean((predicted_bag_trees_train - ur_train$Unemployment_Rate)^2)) #   0.04688094

# RMSE
(rmse_bag_train = sqrt(mse_bag_train))  #  0.2165201

# -------------------------- PERFORMANCE ON TEST DATA ---------------------------------#
# Prediction on Test Data
predicted_bag_trees_test = predict(final_bagged_model, ur_test)

cat("Test Performance Metrics:\n")
# MSE
(mse_bag_test = mean((predicted_bag_trees_test - ur_test$Unemployment_Rate)^2))  # 5.290766

# RMSE
(rmse_bag_test = sqrt(mse_bag_test))  # 2.300167
```

> final_bagged_model

Call:
 randomForest(formula = Unemployment_Rate ~ ., data = ur_train,      mtry = 7, ntree = 134, importance = TRUE, scale = TRUE) 
               Type of random forest: regression
                     Number of trees: 134
No. of variables tried at each split: 7

          Mean of squared residuals: 0.2634143
                    % Var explained: 91.75
> final_bagged_model$importance

                                   %IncMSE IncNodePurity
Percentage_change_GDP          0.010018927      6.535486
Percentage_change_GFCE         0.001444173      4.501828
Percentage_change_FCEAI       -0.005118712      2.734659
Term_of_TI                     0.010328915      5.168316
Consumer_Price_Index           1.249294610    122.756478
Job_Vacancies                  1.867548779    169.813098
Estimated_Resident_Population  1.816775862    182.325098

> cat("Training Performance Metrics:\n")
Training Performance Metrics:
> # MSE
> (mse_bag_train = mean((predicted_bag_trees_train - ur_train$Unemployment_Rate)^2)) #   0.04688094
[1] 0.04688094

> # RMSE
> (rmse_bag_train = sqrt(mse_bag_train))  #  0.2165201
[1] 0.2165201

> cat("Test Performance Metrics:\n")
Test Performance Metrics:
> # MSE
> (mse_bag_test = mean((predicted_bag_trees_test - ur_test$Unemployment_Rate)^2))  # 5.290766
[1] 5.290766

> # RMSE
> (rmse_bag_test = sqrt(mse_bag_test))  # 2.300167
[1] 2.300167

# --------------------------------------- RANDOM FOREST -----------------------------------#

## USING THE RANDOM FORREST FUNCTION
```{r}
# Data Partitioning
set.seed(99) # Set the seed for model consistency

# Training Data = Dec 1982 to Dec 2020
ur_training_filter = (unemployment_rate$Year < 2021) | (unemployment_rate$Year == 2020 & unemployment_rate$Month == "Dec")
# Test Data = Jan 2021 to March 2023
ur_test_filter = (unemployment_rate$Year >= 2021) | (unemployment_rate$Year == 2021 & unemployment_rate$Month %in% c("Mar"))

# Create the training and test sets
ur_train = unemployment_rate[ur_training_filter, ]
ur_test = unemployment_rate[ur_test_filter, ]

# Remove Year, Month, and Quarter since they are not part of the original predictors
ur_train = ur_train[, -c(2, 3, 4)]
ur_test = ur_test[, -c(2, 3, 4)]

#--------------------------------------------------------------------------------------#

start_time_9 = Sys.time()  # Start the timer
set.seed(99) # Set the seed for model consistency

# Hyperparameters for Random Forest
hyperparameters = expand.grid(
  ntree = c(20:500),                # Number of trees in the forest
  max_depth = c(5, 10, 15),         # Maximum depth of each tree
  mtry = 3                      # Number of features considered for splitting at each node ( 7/3 ~ here is 3)
  )

# Initialize variables to store the best MSE and corresponding hyperparameters
best_mse_train_rf = Inf
best_hyperparameters_train = NULL

# Random Forest loop
for (i in 1:nrow(hyperparameters)) 
  {

    # Train a random forest regression model with the current hyperparameters
    rf_ur = randomForest(Unemployment_Rate ~ .,
                         data = ur_train,
                         ntree = hyperparameters$ntree[i],
                         max_depth = hyperparameters$max_depth[i],
                         mtry = hyperparameters$mtry[i],
                         importance = TRUE)
  
    # Make predictions on the Training set
    predictions_train = predict(rf_ur, newdata = ur_train)
  
    # Calculate MSE for the training set
    mse_train_rf = mean((predictions_train - ur_train$Unemployment_Rate)^2)
  
    # Check if this model has a better MSE compared to the overall best
    if (mse_train_rf < best_mse_train_rf) 
      {
        best_mse_train_rf = mse_train_rf
        best_hyperparameters_train = hyperparameters$ntree[i]  # Update the best hyperparameters
        best_model_rf_train = rf_ur  # Update the best model
      }
}

#start_time_9 = Sys.time()  # Start the timer
end_time_9 = Sys.time()  # End the timer
(time_diff_9 = end_time_9 - start_time_9) # Time difference of 2.279682 mins


# # Calculate RMSE for training data
# (rmse_train_rf = sqrt(best_mse_train_rf))
# 
# # Print the best MSE for the training set
# cat("Best Mean Squared Error (MSE) for Training Data:", best_mse_train_rf, "\n")
# 
# # Print the RMSE for the training set
# cat("Root Mean Squared Error (RMSE) for Training Data:", rmse_train_rf, "\n")

# Print the best hyperparameters for the training set
cat("Best Hyperparameters:\n",
    "Number of Trees (ntree):", best_hyperparameters_train, "\n",
    "Maximum Depth:", hyperparameters$max_depth[which.min(best_mse_train_rf)], "\n",
    "mtry:", hyperparameters$mtry[which.min(best_mse_train_rf)], "\n"
    )

print(best_model_rf_train) # Summary of the best model
best_model_rf_train$importance # Variable Importance for Random Forest
varImpPlot(best_model_rf_train, main = "Variable Importance - Random Forest") # Variable Importance Plot

```
#(time_diff_9 = end_time_9 - start_time_9) # Time difference of 3.565514 mins

Best Hyperparameters:
 Number of Trees (ntree): 68 
 Maximum Depth: 5 
 mtry: 3 
> 
> print(best_model_rf_train) # Summary of the best model

Call:
 randomForest(formula = Unemployment_Rate ~ ., data = ur_train,      ntree = hyperparameters$ntree[i], max_depth = hyperparameters$max_depth[i],      mtry = hyperparameters$mtry[i], importance = TRUE) 
               Type of random forest: regression
                     Number of trees: 68
No. of variables tried at each split: 3

          Mean of squared residuals: 0.1963705
                    % Var explained: 93.85
                    
> best_model_rf_train$importance # Variable Importance for Random Forest
                                   %IncMSE IncNodePurity
Percentage_change_GDP          0.041499850      7.418429
Percentage_change_GFCE        -0.005102325      8.077833
Percentage_change_FCEAI        0.012504546      5.893391
Term_of_TI                     0.039759585      8.973473
Consumer_Price_Index           2.160515622    185.569760
Job_Vacancies                  1.892137590    145.922196
Estimated_Resident_Population  1.626099461    134.686699

# Build The model using the Best Hyperparameters for Random Forest

```{r}
start_time_10 = Sys.time()  # Start the timer
set.seed(99) # Set seed for model consistency

# Train the final model with the best hyperparameters on the full training data
final_rf_model = randomForest(Unemployment_Rate ~. ,data = ur_train, 
                              mtry = 3,
                              ntree = 68,
                              max_depth = 5,
                              importance = TRUE) # using all 7/3 predictors

end_time_10 = Sys.time()  # End the timer
(time_diff_10 = end_time_10 - start_time_10) # Time difference of 0.02725506 secs


final_rf_model # Model Results
final_rf_model$importance # Variable Importance

# Variable Importance Plot
varImpPlot(final_rf_model, main = "Variable Importance - Random Forest with the Optimal Hyperparameters")

# -------------------------- PERFORMANCE ON TRAINING DATA ---------------------------------#
# Prediction on Training Data
predicted_rf_trees_train = predict(final_rf_model, ur_train)

cat("Training Performance Metrics:\n")
# MSE
(mse_rf_train = mean((predicted_rf_trees_train - ur_train$Unemployment_Rate)^2)) # 0.04358691

# RMSE
(rmse_rf_train = sqrt(mse_rf_train))  #  0.2087748

# -------------------------- PERFORMANCE ON TEST DATA -------------------------------
# Prediction on Test Data
predicted_rf_trees_test = predict(final_rf_model, ur_test)

cat("Test Performance Metrics:\n")
# MSE
(mse_rf_test = mean((predicted_rf_trees_test - ur_test$Unemployment_Rate)^2)) # 3.776315

# RMSE
(rmse_rf_test = sqrt(mse_rf_test))  #  1.943274
```

> final_rf_model

Call:
 randomForest(formula = Unemployment_Rate ~ ., data = ur_train,      mtry = 3, ntree = 68, max_depth = 5, split_criterion = "gini",      importance = TRUE, ) 
               Type of random forest: regression
                     Number of trees: 68
No. of variables tried at each split: 3

          Mean of squared residuals: 0.237437
                    % Var explained: 92.57
> final_rf_model$importance # Important Variables 
                                  %IncMSE IncNodePurity
Percentage_change_GDP          0.02663089      7.196208
Percentage_change_GFCE        -0.01131663      6.927658
Percentage_change_FCEAI        0.01610534      8.258370
Term_of_TI                     0.02125517     13.497688
Consumer_Price_Index           1.38698353    152.640650
Job_Vacancies                  1.65711359    143.448320
Estimated_Resident_Population  1.76182371    161.676639

> # -------------------------- PERFORMANCE ON TRAINING DATA ---------------------------------#
> # Prediction on Training Data
> predicted_rf_trees_train = predict(final_rf_model, ur_train)
> 
> cat("Training Performance Metrics:\n")
Training Performance Metrics:
> # MSE
> (mse_rf_train = mean((predicted_rf_trees_train - ur_train$Unemployment_Rate)^2)) 
[1] 0.04358691
> 
> # RMSE
> (rmse_rf_train = sqrt(mse_rf_train))  
[1] 0.2087748
> 
> # -------------------------- PERFORMANCE ON TEST DATA -------------------------------
> # Prediction on Test Data
> predicted_rf_trees_test = predict(final_rf_model, ur_test)
> 
> cat("Test Performance Metrics:\n")
Test Performance Metrics:
> # MSE
> (mse_rf_test = mean((predicted_rf_trees_test - ur_test$Unemployment_Rate)^2)) 
[1] 3.776315
> 
> # RMSE
> (rmse_rf_test = sqrt(mse_rf_test))  
[1] 1.943274

# ----------------------------------------- BOOSTING -------------------------------------------------------------
# Gradient Booosting

```{r}

start_time_11 = Sys.time()  # Start the timer
set.seed(99) # Set seed for model consistency

# 10 -fold CV on training data
ur_train_control = trainControl(method = "cv",
                                number = 10,
                                savePredictions = "final") # save

# Build the model on trainig data
boost_ur = caret::train(Unemployment_Rate ~ . , data = ur_train,
                         method = "gbm",
                         trControl = ur_train_control)

#start_time_11 = Sys.time()  # Start the timer
end_time_11 = Sys.time()  # End the timer
(time_diff_11 = end_time_11 - start_time_11) # Time difference of 0.8710001 secs

boost_ur # Model results
summary(boost_ur) # Model summary

boost_ur$bestTune # Best Model

plot(boost_ur, main = "Boosting using 10 -fold CV") # Plot the model

# Prediction on Test Data
predicted_xgb = predict(boost_ur, ur_test)

# MSE
(mse_sgb_boost = mean((predicted_xgb - ur_test$Unemployment_Rate)^2))   # 2.418606

# RMSE
(rmse_sgb_boost = sqrt(mse_sgb_boost)) # 1.555187

```
> summary(boost_ur)
                                                        var   rel.inf
Job_Vacancies                                 Job_Vacancies 44.193020
Consumer_Price_Index                   Consumer_Price_Index 42.677890
Estimated_Resident_Population Estimated_Resident_Population  3.965492
Percentage_change_GDP                 Percentage_change_GDP  3.508895
Term_of_TI                                       Term_of_TI  2.402376
Percentage_change_FCEAI             Percentage_change_FCEAI  1.712144
Percentage_change_GFCE               Percentage_change_GFCE  1.540181

> plot(boost_ur, main = "Boosting") # Plot the model

> boost_ur$bestTune # Best Model

  n.trees interaction.depth shrinkage n.minobsinnode
8     100                 3       0.1             10

> plot(boost_ur, main = "Boosting") # Plot the model

> # Prediction on Test Data
> predicted_xgb = predict(boost_ur, ur_test)

> # MSE
> (mse_sgb_boost = mean((predicted_xgb - ur_test$Unemployment_Rate)^2))   # 2.418606
[1] 2.418606
> 
> # RMSE
> (rmse_sgb_boost = sqrt(mse_sgb_boost)) # 1.555187
[1] 1.555187

# TUNE THE BOOSTING PARAMETERS
```{r}
# Set the training control for Boosting
ur_train_control = trainControl(method = "cv",
                                number = 10,
                                savePredictions = "final"
                                 ) # save

# Set the hyperparameters for Boosting
gbmGrid =  expand.grid(interaction.depth = c(2,5,10,15),  # Max Tree Depth
                        n.trees = c(1:150), # Boosting Iterations
                        shrinkage = c(0.001, 0.1,0.15),# Shrinkage = Learning Rate
                        n.minobsinnode = c(5,10,20) # minimum number of training set samples in a node to commence splitting 
                       )

start_time_12 = Sys.time()  # Start the timer
set.seed(99) # Set seed for Model Consistency
boost_ur_tune = caret::train(Unemployment_Rate ~ . , data = ur_train,
                         method = "gbm",
                         trControl = ur_train_control,
                         tuneGrid = gbmGrid)

#start_time_12 = Sys.time()  # Start the timer
end_time_12 = Sys.time()  # End the timer
(time_diff_12 = end_time_12 - start_time_12) # Time difference of 3.900011 mins

summary(boost_ur_tune) # Model Summary 
boost_ur_tune$bestTune # Best Parameters

plot(boost_ur_tune , metric = "RMSE" , main = "Hyperparameter Importance - Boosting") # Plot the model

# # Prediction on Test Data
# predicted_xgb_tuned = predict(boost_ur_tune, ur_test)
# 
# # MSE
# (mse_sgb_boost_tuned = mean((predicted_xgb_tuned - ur_test$Unemployment_Rate)^2)) 
# 
# # RMSE
# (rmse_sgb_boost_tuned = sqrt(mse_sgb_boost_tuned)) 

```
> summary(boost_ur_tune)
                                                        var   rel.inf
Consumer_Price_Index                   Consumer_Price_Index 51.658775
Job_Vacancies                                 Job_Vacancies 34.739494
Percentage_change_GDP                 Percentage_change_GDP  3.993640
Term_of_TI                                       Term_of_TI  2.863609
Percentage_change_GFCE               Percentage_change_GFCE  2.511384
Estimated_Resident_Population Estimated_Resident_Population  2.148509
Percentage_change_FCEAI             Percentage_change_FCEAI  2.084589

> boost_ur_tune$bestTune
     n.trees interaction.depth shrinkage n.minobsinnode
4713      63                10      0.15             10

> # MSE
> (mse_sgb_boost_tuned = mean((predicted_xgb_tuned - ur_test$Unemployment_Rate)^2)) # 2.516795 
[1] 2.516795
> 
> # RMSE
> (rmse_sgb_boost_tuned = sqrt(mse_sgb_boost_tuned)) # 1.611611
[1] 1.586441

# BOOSTING - Using the Best HyperParameters
```{r}
#start_time_13 = Sys.time()  # Start the timer
set.seed(99) # Set seed for model consistency
ur_train_control = trainControl(method = "cv",
                                number = 10,
                                savePredictions = "final",
                                 ) # save

gbmGrid_2 =  expand.grid(interaction.depth = 10,  # Max Tree Depth
                        n.trees =  63, # Boosting Iterations
                        shrinkage = 0.15 ,# Shrinkage = Learning Rate
                        n.minobsinnode = 10 # minimum number of training set samples in a node to commence splitting 
                       )

start_time_13 = Sys.time()  # Start the timer
set.seed(99) # Set seed for model consistency
boost_best_ur = caret::train(Unemployment_Rate ~ . , data = ur_train,
                 method = "gbm",
                 trControl = ur_train_control,
                 tuneGrid = gbmGrid_2)  

#start_time_13 = Sys.time()  # Start the timer
end_time_13 = Sys.time()  # End the timer
(time_diff_13 = end_time_13 - start_time_13) # Time difference of 0.4310961 secs


boost_best_ur # Model results
summary(boost_best_ur) # Model summary

boost_best_ur$bestTune # Best Model

# -------------------------- PERFORMANCE ON TRAINING DATA ---------------------------------#
# Prediction on Training Data
predicted_sgb_train = predict(boost_best_ur, ur_train)

cat("Train Performance Metrics:\n")
# MSE
(mse_sgb_train = mean((predicted_sgb_train - ur_train$Unemployment_Rate)^2))   # 0.1428478

# RMSE
(rmse_sgb_train = sqrt(mse_sgb_train))  # 0.3779522

# -------------------------- PERFORMANCE ON TEST DATA ---------------------------------#
# Prediction on Test Data
predicted_sgb_test = predict(boost_best_ur, ur_test)

cat("Test Performance Metrics:\n")
# MSE
(mse_sgb_test = mean((predicted_sgb_test - ur_test$Unemployment_Rate)^2))   # 2.460212

# RMSE
(rmse_sgb_test = sqrt(mse_sgb_test)) # 1.568506

```
> boost_best_ur
Stochastic Gradient Boosting 

156 samples
  7 predictor

No pre-processing
Resampling: Cross-Validated (10 fold) 
Summary of sample sizes: 141, 140, 140, 140, 140, 140, ... 
Resampling results:

  RMSE       Rsquared  MAE      
  0.6361827  0.887461  0.4326512

Tuning parameter 'n.trees' was held constant at a value of 63
Tuning parameter 'interaction.depth' was held constant at a value of
 10
Tuning parameter 'shrinkage' was held constant at a value of 0.15
Tuning parameter 'n.minobsinnode' was held constant at a value
 of 10
 
> summary(boost_best_ur)
                                                        var   rel.inf
Consumer_Price_Index                   Consumer_Price_Index 49.368438
Job_Vacancies                                 Job_Vacancies 38.649790
Estimated_Resident_Population Estimated_Resident_Population  3.550039
Percentage_change_GDP                 Percentage_change_GDP  3.177617
Percentage_change_GFCE               Percentage_change_GFCE  2.004011
Term_of_TI                                       Term_of_TI  1.934238
Percentage_change_FCEAI             Percentage_change_FCEAI  1.315867


> boost_best_ur$bestTune # Best Model

  n.trees interaction.depth shrinkage n.minobsinnode
1      63                10      0.15             10

> cat("Train Performance Metrics:\n")
Train Performance Metrics:
> # MSE
> (mse_sgb_train = mean((predicted_sgb_train - ur_train$Unemployment_Rate)^2))  
[1] 0.1428478
> 
> # RMSE
> (rmse_sgb_train = sqrt(mse_sgb_train)) 
[1] 0.3779522
> 
> # -------------------------- PERFORMANCE ON TEST DATA ---------------------------------#
> # Prediction on Test Data
> predicted_sgb_test = predict(boost_best_ur, ur_test)

> cat("Test Performance Metrics:\n")
Test Performance Metrics:
> # MSE
> (mse_sgb_test = mean((predicted_sgb_best - ur_test$Unemployment_Rate)^2))   # 2.460212
[1] 2.460212
> 
> # RMSE
> (rmse_sgb_test = sqrt(mse_sgb_test)) # 1.568506
[1] 1.568506

# ----------------------------------------- IV.SUPPORT VECTOR MACHINES -----------------------------------------------
# IV. SUPPORT VECTOR MACHINES

# 1.  With kernel = "linear" and hyperparameters : cost and gamma
```{r}
# With kernel = "linear" with cost and gamma

start_time_14 = Sys.time()  # Start the timer
set.seed(99) # Set seed for model consistency
tune_ur_linear = tune(svm,
                 Unemployment_Rate ~. ,
                 data = ur_train, 
                 scale = TRUE,
                 kernel= c("linear"), 
                 ranges=list(cost= c(0.001, 0.01, 0.1, 1,5,10,50),
                             gamma = c(0.001, 0.01, 0.1, 1,5,10,50)))

# start_time_14 = Sys.time()  # Start the timer
end_time_14 = Sys.time()  # End the timer
(time_diff_14 = end_time_14 - start_time_14) # Time difference of 9.060912 secs

tune_ur_linear  # Model results
summary(tune_ur_linear$best.model) # Summary of the best model

plot(tune_ur_linear) # Plot the model of SVM

tune_ur_linear$best.model # Best Model
tune_ur_linear$best.parameters  # Best Parameters 
tune_ur_linear$best.performance # Best Performance

# -------------------------- PERFORMANCE ON TRAIN DATA ---------------------------------#
# Predictions on Train Data using the best model
svm_predict_linear_train = predict(tune_ur_linear$best.model, ur_train)

# MSE
(mse_svm_linear_train = mean((svm_predict_linear_train - ur_train$Unemployment_Rate)^2)) # 0.9088124

# RMSE
(rmse_svm_linear_train = sqrt(mse_svm_linear_train))  # 0.9533165

# -------------------------- PERFORMANCE ON TEST DATA ---------------------------------#
# Predictions on Test Data using the best model
svm_predict_linear_test = predict(tune_ur_linear$best.model, ur_test)

# MSE
(mse_svm_linear_test = mean((svm_predict_linear_test - ur_test$Unemployment_Rate)^2)) # 31.71556

# RMSE
(rmse_svm_linear_test = sqrt(mse_svm_linear_test))  # 5.631657
```
> tune_ur_linear

Parameter tuning of svm:

- sampling method: 10-fold cross validation 

- best parameters:
 cost gamma
    1 0.001

- best performance: 1.156467 

> summary(tune_ur_linear$best.model)

> tune_ur_linear$best.model

Call:
best.tune(method = svm, train.x = Unemployment_Rate ~ ., data = ur_train, ranges = list(cost = c(0.001, 0.01, 
    0.1, 1, 5, 10, 50), gamma = c(0.001, 0.01, 0.1, 1, 5, 10, 50)), scale = TRUE, kernel = c("linear"))


Parameters:
   SVM-Type:  eps-regression 
 SVM-Kernel:  linear 
       cost:  1 
      gamma:  0.001 
    epsilon:  0.1 


Number of Support Vectors:  121

> tune_ur_linear$best.performance

[1] 1.156467

> # -------------------------- PERFORMANCE ON TRAIN DATA ---------------------------------#
> # Predictions on Train Data using the best model
> svm_predict_linear_train = predict(tune_ur_linear$best.model, ur_train)
> 
> # MSE
> (mse_svm_linear_train = mean((svm_predict_linear_train - ur_train$Unemployment_Rate)^2)) # 0.9088124
[1] 0.9088124
> 
> # RMSE
> (rmse_svm_linear_train = sqrt(mse_svm_linear_train))  # 0.9533165
[1] 0.9533165
> 
> # -------------------------- PERFORMANCE ON TEST DATA ---------------------------------#
> # Predictions on Test Data using the best model
> svm_predict_linear_test = predict(tune_ur_linear$best.model, ur_test)
> 
> # MSE
> (mse_svm_linear_test = mean((svm_predict_linear_test - ur_test$Unemployment_Rate)^2)) # 31.71556
[1] 31.71556
> 
> # RMSE
> (rmse_svm_linear_test = sqrt(mse_svm_linear_test))  # 5.631657
[1] 5.631657


# ---------------------------------------------------------------------------------------

# 2.  With kernel = "radial" and hyperparameters : cost and gamma
```{r}
# With kernel = "linear" with cost and gamma

start_time_15 = Sys.time()  # Start the timer
set.seed(99)
tune_ur_radial = tune(svm,
                 Unemployment_Rate ~. ,
                 data = ur_train, 
                 scale = TRUE,
                 kernel= "radial", 
                 ranges=list(cost= c(0.001, 0.01, 0.1, 1,5,10,50),
                             gamma = c(0.001, 0.01, 0.1, 1,5,10,50)))

# start_time_15 = Sys.time()  # Start the timer
end_time_15 = Sys.time()  # End the timer
(time_diff_15 = end_time_15 - start_time_15) # Time difference of 4.546159 secs

tune_ur_radial # Model results
summary(tune_ur_radial$best.model) # Summary of the best model
plot(tune_ur_radial) # Plot the model

tune_ur_radial$best.model # Best Model
tune_ur_radial$best.parameters  # Best Parameters
tune_ur_radial$best.performance # Best Performance



# -------------------------- PERFORMANCE ON TRAIN DATA ---------------------------------#
# Predictions on Train Data using the best model
svm_predict_radial_train = predict(tune_ur_radial$best.model, ur_train)

# MSE
(mse_svm_radial_train = mean((svm_predict_radial_train - ur_train$Unemployment_Rate)^2)) # 0.2476974

# RMSE
(rmse_svm_radial_train = sqrt(mse_svm_radial_train))  # 0.4976921

# -------------------------- PERFORMANCE ON TEST DATA ---------------------------------#

# Predictions on the Test Data using the Best Model
svm_predict_radial_test = predict(tune_ur_radial$best.model, ur_test)

# MSE
(mse_svm_radial_test = mean((svm_predict_radial_test - ur_test$Unemployment_Rate)^2)) # 2.705984

# RMSE
(rmse_svm_radial_test = sqrt(mse_svm_radial_test))  # 1.644988
```

> tune_ur_radial$best.model

Call:
best.tune(method = svm, train.x = Unemployment_Rate ~ ., data = ur_train, ranges = list(cost = c(0.001, 0.01, 
    0.1, 1, 5, 10, 50), gamma = c(0.001, 0.01, 0.1, 1, 5, 10, 50)), scale = TRUE, kernel = "radial")


Parameters:
   SVM-Type:  eps-regression 
 SVM-Kernel:  radial 
       cost:  5 
      gamma:  0.1 
    epsilon:  0.1 


Number of Support Vectors:  111

> tune_ur_radial$best.parameters 
   cost gamma
19    5   0.1

> # -------------------------- PERFORMANCE ON TRAIN DATA ---------------------------------#
> # Predictions on Train Data using the best model
> svm_predict_radial_train = predict(tune_ur_radial$best.model, ur_train)
> 
> # MSE
> (mse_svm_radial_train = mean((svm_predict_radial_train - ur_train$Unemployment_Rate)^2)) # 0.2476974
[1] 0.2476974
> 
> # RMSE
> (rmse_svm_radial_train = sqrt(mse_svm_radial_train))  # 0.4976921
[1] 0.4976921
> 
> # -------------------------- PERFORMANCE ON TEST DATA ---------------------------------#
> 
> # Predictions on the Test Data using the Best Model
> svm_predict_radial_test = predict(tune_ur_radial$best.model, ur_test)
> 
> # MSE
> (mse_svm_radial_test = mean((svm_predict_radial_test - ur_test$Unemployment_Rate)^2)) # 2.705984
[1] 2.705984
> 
> # RMSE
> (rmse_svm_radial_test = sqrt(mse_svm_radial_test))  # 1.644988
[1] 1.644988

# 3.  With kernel = "polynomial" and hyperparameters : cost and gamma
```{r}
# With kernel = "linear" with cost and gamma
start_time_16 = Sys.time()  # Start the timer
set.seed(99) # Set seed for model consistency
tune_ur_poly = tune(svm,
                 Unemployment_Rate ~. ,
                 data = ur_train, 
                 scale = TRUE,
                 kernel= "polynomial", 
                 ranges=list(cost= c(0.001, 0.01, 0.1, 1,5),
                             gamma = c(0.001, 0.01, 0.1, 1),
                             degree = c(2,3,4)
                             ))

# start_time_16 = Sys.time()  # Start the timer
end_time_16 = Sys.time()  # End the timer
(time_diff_16 = end_time_16 - start_time_16) # Time difference of 36.7868 secs

tune_ur_poly # Model results
summary(tune_ur_poly$best.model) # Summary of the best model

#plot(tune_ur_poly)

tune_ur_poly$best.model # Best model 
tune_ur_poly$best.parameters # Best Parameters
tune_ur_poly$best.performance # Best Performance

# -------------------------- PERFORMANCE ON TRAIN DATA ---------------------------------#
# Predictions on Train Data using the best model
svm_predict_poly_train = predict(tune_ur_poly$best.model, ur_train)

# MSE
(mse_svm_poly_train = mean((svm_predict_poly_train - ur_train$Unemployment_Rate)^2)) # 2.733716

# RMSE
(rmse_svm_poly_train = sqrt(mse_svm_poly_train))  # 1.653395

# -------------------------- PERFORMANCE ON TEST DATA ---------------------------------#
# Predictions on the Test Data using the Best Model
svm_predict_poly_test = predict(tune_ur_poly$best.model, ur_test)

# MSE
(mse_svm_poly_test = mean((svm_predict_poly_test - ur_test$Unemployment_Rate)^2)) # 0.5671939

# RMSE
(rmse_svm_poly_test = sqrt(mse_svm_poly_test))  # 0.7531228

```
> tune_ur_poly$best.model

Call:
best.tune(method = svm, train.x = Unemployment_Rate ~ ., data = ur_train, ranges = list(cost = c(0.001, 0.01, 0.1, 
    1, 5), gamma = c(0.001, 0.01, 0.1, 1), degree = c(2, 3, 4)), scale = TRUE, kernel = "polynomial")


Parameters:
   SVM-Type:  eps-regression 
 SVM-Kernel:  polynomial 
       cost:  1 
     degree:  2 
      gamma:  0.1 
     coef.0:  0 
    epsilon:  0.1 


Number of Support Vectors:  137

> tune_ur_poly$best.parameters 
   cost gamma degree
14    1   0.1      2
> tune_ur_poly$best.performance
[1] 3.185752
> 
> # Predictions on the Test Data using the Best Model
> svm_predict_poly = predict(tune_ur_poly$best.model, ur_test)
> 
> (mse_svm_poly = mean((svm_predict_poly - ur_test$Unemployment_Rate)^2)) # [1] 2.705984
[1] 0.5671939
> 
> # RMSE
> (rmse_svm_poly = sqrt(mse_svm_poly))  # 0.7531228
[1] 0.7531228

# Build the Model with the Best Hyperparameters ( SVM with kernel = polynomial)
```{r}

start_time_17 = Sys.time()  # Start the timer
set.seed(99)
svm_ur_best = svm(Unemployment_Rate ~. , data = ur_train , 
                  kernel = "poly", 
                  scale = TRUE,
                  cost = 1,
                  gamma= 0.1,
                  degree = 2) 

# start_time_17 = Sys.time()  # Start the timer
end_time_17 = Sys.time()  # End the timer
(time_diff_17 = end_time_17 - start_time_17) # Time difference of 0.01179504 secs

svm_ur_best
summary(svm_ur_best)


# -------------------------- PERFORMANCE ON TEST DATA ---------------------------------#
# Predictions on the Test Data using the Best Model
svm_predict_poly_test_2 = predict(svm_ur_best, ur_test)

# MSE
(mse_svm_poly_test_2 = mean((svm_predict_poly_test_2 - ur_test$Unemployment_Rate)^2)) # 0.5671939

# RMSE
(rmse_svm_poly_test_2 = sqrt(mse_svm_poly_test_2))  # 0.75312
```


# --------------------------- ALL ML MODEL RESULTS  ------------------------------------------

```{r}
# MSE results all models
# --------------------------------MSE train data 
mse_ridge_train
mse_lasso_train
mse_cart_train
mse_bag_train
mse_rf_train
mse_sgb_train
mse_svm_linear_train
mse_svm_radial_train
mse_svm_poly_train


#-------------------------------- RMSE train data
rmse_ridge_train
rmse_lasso_train
rmse_cart_train
rmse_bag_train
rmse_rf_train
rmse_sgb_train
rmse_svm_linear_train
rmse_svm_radial_train
rmse_svm_poly_train

#--------------------------------- MSE test Data
mse_ridge_test
mse_lasso_test
mse_cart_test
mse_bag_test
mse_rf_test
mse_sgb_test
mse_svm_linear_test
mse_svm_radial_test
mse_svm_poly_test

#--------------------------------- RMSE test Data
rmse_ridge_test
rmse_lasso_test
rmse_cart_test
rmse_bag_test
rmse_rf_test
rmse_sgb_test
rmse_svm_linear_test
rmse_svm_radial_test
rmse_svm_poly_test


MSE_TRAIN = c(mse_ridge_train, mse_lasso_train, mse_cart_train , mse_bag_train , mse_rf_train, mse_sgb_train, mse_svm_linear_train,mse_svm_radial_train,mse_svm_poly_train)

RMSE_TRAIN = c(rmse_ridge_train, rmse_lasso_train, rmse_cart_train,rmse_bag_train, rmse_rf_train, rmse_sgb_train,rmse_svm_linear_train,rmse_svm_radial_train,rmse_svm_poly_train)

MSE_TEST = c(mse_ridge_test, mse_lasso_test, mse_cart_test , mse_bag_test , mse_rf_test, mse_sgb_test, mse_svm_linear_test,mse_svm_radial_test,mse_svm_poly_test)

RMSE_TEST = c(rmse_ridge_test, rmse_lasso_test, rmse_cart_test,rmse_bag_test, rmse_rf_test, rmse_sgb_test ,rmse_svm_linear_test,rmse_svm_radial_test,rmse_svm_poly_test)

Models = c("RIDGE REGRESSION","LASSO REGRESSION","CART","BAGGING","RANDOM FOREST", "Stochastic Gradient Boosting", "SVM - Linear", "SVM - Radial", "SVM - Polynomial")


ML_results = data.frame(Models,MSE_TRAIN,RMSE_TRAIN,MSE_TEST,RMSE_TEST)
View(ML_results)

# Find the minimum MSE and RMSE for training data
(min_MSE_train = min(ML_results$MSE_TRAIN)) # Random Forest
(min_RMSE_train = min(ML_results$RMSE_TRAIN)) # Random Forest

# Find the minimum MSE and RMSE for test data
(min_MSE_test = min(ML_results$MSE_TEST)) # SVM Polynomial
(min_RMSE_test = min(ML_results$RMSE_TEST)) # SVM Polynomial
```




# -------------------------- ARTIFICIAL NEURAL NETWORKS ------------------------------------#
# Neural Network Libraries
```{r}
# # Neural Network Library
# # library(reticulate)
# #install.packages("keras")
# #library(keras)
# #library(remotes)
# remotes::install_github(paste0("rstudio/", c("reticulate", "tensorflow", "keras")))
# reticulate::install_python()
# install_tensorflow(envname = "r-tensorflow")
# library(tensorflow)
```

# Data Partitioning
```{r}
set.seed(99) # Set the seed for model consistency

# Training Data = Dec 1982 to Dec 2020
ur_training_filter = (unemployment_rate$Year < 2021) | (unemployment_rate$Year == 2020 & unemployment_rate$Month == "Dec")
# Test Data = Jan 2021 to March 2023
ur_test_filter = (unemployment_rate$Year >= 2021) | (unemployment_rate$Year == 2021 & unemployment_rate$Month %in% c("Mar"))

# Create the training and test sets
ur_train = unemployment_rate[ur_training_filter, ]
ur_test = unemployment_rate[ur_test_filter, ]

# Remove Year, Month, and Quarter since they are not part of the original predictors
ur_train = ur_train[, -c(2, 3, 4)]
ur_test = ur_test[, -c(2, 3, 4)]

# Seperate the response variables for train and test
ur_train_nn_y = ur_train$Unemployment_Rate # Train response
ur_test_nn_y = ur_test$Unemployment_Rate # Test response

# Seperate predictor variables for train and test
ur_train_nn_x = ur_train[,-1] # Train predictors
ur_test_nn_x = ur_test[-1] # Test predictors

# Scale the Data
ur_train_nn_x_scaled = scale(ur_train_nn_x) # Scaled Predictor variables for Training Data
ur_test_nn_x_scaled = scale(ur_test_nn_x) # Scaled Predictor variables for Training Data
summary(ur_train_nn_x_scaled) # Check summary of the scaled data
summary(ur_test_nn_x_scaled) # Check summary of test data

dim(ur_train_nn_x_scaled)  # 156  x  7
dim(ur_test_nn_x_scaled) # 10 x 7
```


```{r}
# NN with adam

# Define the neural network model
nn_model_ur = keras_model_sequential() %>%
  layer_dense(units = 64, activation = "relu", input_shape = c(ncol(ur_test_nn_x_scaled))) %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 1)

# Define metrics
nn_model_ur %>% compile(optimizer = "adam",
                          loss = "mse",
                          metric = c("mae"))


# Create model history
nn_model_ur_adam = nn_model_ur %>% fit(ur_train_nn_x_scaled, 
                                               ur_train_nn_y,
                                               epochs = 300,
                                               batch_size = 8,
                                               validation_split = 1/3) # Random fraction of the training data to be with-held during each epoch to be used as a pseudo validation data set and the remaining fraction of the training data are used as to train the network parameters during the epoch.


# A plot of the bootstrapped validation MAE loss from network model training history.
nn_epoch_plot_adam = data.frame(x = c(5:nn_model_ur_adam$params$epochs), 
                                y = nn_model_ur_adam$metrics$val_mae[-c(1:4)])

# Visualisation for the nn plots
ggplot(nn_epoch_plot_adam, aes(x=x, y=y)) +geom_smooth() +xlab("epoch") +ylab("Estimated Validation MAE loss") + geom_vline(xintercept = 78, linetype="dotted", color = "orange", size=1)
```

# nn_model_ur_adam 
Final epoch (plot to see history):
    loss: 0.01979
     mae: 0.09804
val_loss: 3.782
 val_mae: 1.633


#  Build the model based on the Model based on the Lowest error rate ( Epoch value = 78) 
```{r}
#  Build the model based on the Model based on the Lowest error rate ( Epoch value = 78) 
#----Train performance ----

# Define the neural network model
nn_model_ur_train = keras_model_sequential() %>%
  layer_dense(units = 64, activation = "relu", input_shape = c(ncol(ur_train_nn_x_scaled))) %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 1)


# Define metrics
nn_model_ur_train %>% compile(optimizer = "adam",
                          loss = "mse",
                          metric = c("mae"))

# Build the model on Training Data

start_time_18 = Sys.time()  # Start the timer
set.seed(99)
nn_model_ur_train %>% fit(ur_train_nn_x_scaled, 
                      ur_train_nn_y,
                      epochs = 78, # Optimal epoch value derived from ggplot
                      validation_data = list(ur_train_nn_x_scaled,ur_train_nn_y),
                      verbose = 1) 

end_time_18 = Sys.time()  # End the timer
(time_diff_18 = end_time_18 - start_time_18) # Time difference 

# Model Performance on Training Data
# After the network has been tuned to a suitable number of epochs, the validation data can be predicted using the evaluate() function.
nn_model_results_ur_adam_train = nn_model_ur_train %>% evaluate(ur_train_nn_x_scaled , ur_train_nn_y)

nn_model_results_ur_adam_train # Model results (summary)

#----Test performance ----

# Define the neural network model
rm(nn_model_ur_2)
nn_model_ur_2 <- keras_model_sequential() %>%
  layer_dense(units = 64, activation = "relu", input_shape = c(ncol(ur_test_nn_x_scaled))) %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 1)


# Define metrics
nn_model_ur_2 %>% compile(optimizer = "adam",
                          loss = "mse",
                          metric = c("mae"))

start_time_19 = Sys.time()  # Start the timer
set.seed(99)
nn_model_ur_2 %>% fit(ur_train_nn_x_scaled, 
                      ur_train_nn_y,
                      epochs = 78, # Optimal epoch value derived from ggplot
                      validation_data = list(ur_test_nn_x_scaled,ur_test_nn_y),
                      verbose = 1) 
end_time_19 = Sys.time()  # End the timer
(time_diff_19 = end_time_19 - start_time_19) 

# Model Performance on Test Data
# After the network has been tuned to a suitable number of epochs, the validation data can be predicted using the evaluate() function.
nn_model_results_ur_adam_2 = nn_model_ur_2 %>% evaluate(ur_test_nn_x_scaled , ur_test_nn_y)

nn_model_results_ur_adam_2 # Model results (summary)
```
# Model Results:

# TRAIN DATA RESULTS:
loss       mae 
0.5611840 0.5605854

# TEST DATA RESULTS:
nn_model_results_ur_adam_2 # Model results (summary)
     loss       mae 
10.342458  2.632585

Model: "sequential_1152"
_________________________________________
 Layer (type)                                          Output Shape                                     Param #            
===========================================================================================================================
 dense_3596 (Dense)                                    (None, 64)                                       512                
 dense_3595 (Dense)                                    (None, 64)                                       4160               
 dense_3594 (Dense)                                    (None, 1)                                        65                 
===========================================================================================================================
Total params: 4737 (18.50 KB)
Trainable params: 4737 (18.50 KB)
Non-trainable params: 0 (0.00 Byte)
_________________________________________

# -------------------- Hyper parameter tuning -----------------------------------#

# Hyper parametertuning
```{r}
# Define hyperparameters
epochs_list_ = c(70, 80, 90, 100)
units_list_ = c(30, 60, 90)
batch_sizes_ = c(10, 30, 50, 70)

# Create a data frame with all combinations of hyperparameters
hyperparameter_combinations = expand.grid(epochs = epochs_list_, units = units_list_, batch_size = batch_sizes_)

# Initialize variables to store the best results and hyperparameters
best_train_loss = Inf
best_hyperparameters = NULL

start_time_19 = Sys.time()  # Start the timer
set.seed(99)
# Iterate through hyperparameter combinations
for (i in 1:nrow(hyperparameter_combinations)) 
  {
    epochs = hyperparameter_combinations$epochs[i]
    units = hyperparameter_combinations$units[i]
    batch_size = hyperparameter_combinations$batch_size[i]
    
    cat("Training model with epochs =", epochs, "units =", units, "batch size =", batch_size, "\n")
    
    # Define the neural network model
    nn_model= keras_model_sequential() %>%
      layer_dense(units = units, activation = "relu", input_shape = c(ncol(ur_train_nn_x_scaled))) %>%
      layer_dense(units = units, activation = "relu") %>%
      layer_dense(units = units, activation = "relu") %>%
      layer_dense(units = 1)
    
    # Compile the model
    nn_model %>% compile(optimizer = "adam",
                         loss = "mse",
                         metrics = c("mae"))
    
    # Train the model on Training Data
    set.seed(99)
    history = nn_model %>% fit(ur_train_nn_x_scaled, 
                                ur_train_nn_y,
                                epochs = epochs,
                                batch_size = batch_size,
                                validation_data = list(ur_train_nn_x_scaled, ur_train_nn_y),
                                verbose = 1)
    
    # Model Performance on Test Data
    nn_model_results_train = nn_model %>% evaluate(ur_train_nn_x_scaled, ur_train_nn_y)
    cat("Test results (summary):\n")
    print(nn_model_results_train)
    
    # Check if the current test results are better than the best so far
    if (nn_model_results_train[[1]] < best_train_loss) 
      {
        (best_train_loss = nn_model_results_train[[1]])
        (best_hyperparameters = c(epochs, units, batch_size))
      }
}
end_time_19 = Sys.time()  # End the timer
(time_diff_19 = end_time_19 - start_time_19) # Time difference of 12.48 minutes

# Print the best hyperparameter results
cat("Best Hyperparameters:\n")
cat("Epochs:", best_hyperparameters[1], "\n") 
cat("Units:", best_hyperparameters[2], "\n") 
cat("Batch Size:", best_hyperparameters[3], "\n")
cat("Best Train Loss:", best_train_loss,"\n")


```

#Best Hyperparameters:
#> cat("Epochs:", best_hyperparameters[1], "\n")
Epochs: 90 
#> cat("Units:", best_hyperparameters[2], "\n")
Units: 90 
#> cat("Batch Size:", best_hyperparameters[3], "\n")
Batch Size: 10 
#> cat("Best Test Loss:", best_test_loss,"\n")
Best Train Loss: 0.08561801

# Build the model using the Optimal Hyperparameters (epoch = 90, units = 90, batch_size  = 10)
```{r}
#---- Prediction on  test data --- 
#Optimal values: epoch = 90, units = 90, batch_size  = 10.
# Layers
best_nn_model = keras_model_sequential() %>%
  layer_dense(units = 90, activation = "relu", input_shape = c(ncol(ur_test_nn_x_scaled))) %>%
  layer_dense(units = 90, activation = "relu") %>%
  layer_dense(units = 1)

# Define metrics
best_nn_model %>% compile(optimizer = "adam",
                          loss = "mse",
                          metrics = c("mae"))

# Set seed for model consistency
set.seed(99)
start_time_best_nn = Sys.time()  # Start the timer

# Build the model
best_nn_model %>% fit(ur_train_nn_x_scaled,
                 ur_train_nn_y,
                 epochs = 90,
                 batch_size = 10,
                 validation_data = list(ur_test_nn_x_scaled, ur_test_nn_y)) 

end_time_best_nn = Sys.time()  # End the timer
(time_diff_best_nn = end_time_best_nn - start_time_best_nn) # Time difference of 16.3403 secs

# After the network has been tuned , the validation data can be predicted using the evaluate() function.
best_nn_model_result = best_nn_model %>% evaluate(ur_test_nn_x_scaled, ur_test_nn_y)

best_nn_model_result # Model summary

#    MSE      mae 
# 9.271559 2.466631  

#RMSE
(best_nn_model_result_rmse = sqrt(best_nn_model_result[['loss']])) # 3.044923
```

# Model results:

Model: "sequential_1487"
_________________________________________
 Layer (type)                                          Output Shape                                     Param #            
===========================================================================================================================
 dense_4930 (Dense)                                    (None, 90)                                       720                
 dense_4929 (Dense)                                    (None, 90)                                       8190               
 dense_4928 (Dense)                                    (None, 1)                                        91                 
===========================================================================================================================
Total params: 9001 (35.16 KB)
Trainable params: 9001 (35.16 KB)
Non-trainable params: 0 (0.00 Byte)
_________________________________________

#    MSE      mae 
# 9.271559 2.466631  

